---
title: "UKB KFRE multimorbidity code_NDT github_ Adapted from: External validation of the performance of competing risks prediction models: a guide through modern methods -  Cause specific hazard models"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---
  
## Steps
  
The steps taken in this file are:   
1. To develop a competing risks prediction model using the cause specific hazards approach.
2. To assess the performance of the model in terms of calibration, discrimination and overall prediction error. We calculate the apparent (no internal validation) validation and the external validation.
3. To assess the potential clinical utility the model using decision curve analysis.

### Installing and loading packages and import data

The following libraries are used in this file, the code chunk below will a) check whether you already have them installed, b) install them for you if not already present, and c) load the packages into the session.

```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/Prediction_CSC/",
  echo = FALSE
)
```

```{r, wdlib, message=FALSE, warning=FALSE, echo=TRUE}
# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  survival,
  rms,
  cmprsk,
  riskRegression,
  mstate,
  pseudo,
  pec,
  riskRegression,
  plotrix,
  knitr,
  splines,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot,
  dplyr,
  tidymodels
)

# Import data ------------------
load(file="~/R working directory/SCREAM/linear_predictor.RData")
rdata <- linear_predictor
vdata <- readRDS("~/R working directory/UKB/new_ukb.rds")

rdata$event <- ifelse(rdata$death==1, 2, 0)
rdata$event <- ifelse(rdata$outcome==1, 1, rdata$event)
rdata = rdata %>% 
  mutate(time = (futime_months/12)) %>% 
  mutate(status= case_when(
    event==2 ~ "death",
    event==1 ~"KF",
    event==0 ~ "cens")) %>% 
  mutate(acr2 = ifelse(acr2 == 0, 0.00001, acr2)) %>% 
  mutate(lnacr2 = (log(acr2)))

vdata$event <- ifelse(vdata$death==1, 2, 0)
vdata$event <- ifelse(vdata$outcome==1, 1, vdata$event)
vdata = vdata %>% 
  mutate(time = (futime_months/12)) %>% 
  mutate(status= case_when(
    event==2 ~ "death",
    event==1 ~"KF",
    event==0 ~ "cens")) %>% 
  mutate(acr2 = ifelse(acr2 == 0, 0.00001, acr2)) %>% 
  mutate(lnacr2 = (log(acr2)))

rdata <- rdata %>%
  mutate(status = factor(case_when(
    event == 2 ~ "death",
    event == 1 ~ "KF",
    event == 0 ~ "cens"
  ), levels = c("death", "KF", "cens")))

vdata <- vdata %>%
  mutate(status = factor(case_when(
    event == 2 ~ "death",
    event == 1 ~ "KF",
    event == 0 ~ "cens"
  ), levels = c("death", "KF", "cens")))

rdata <- rdata %>%
  mutate(sex = factor(case_when(
    female == 0 ~ "female",
    female == 1 ~ "male"
  ), levels = c("female", "male")))


vdata <- vdata %>%
  mutate(sex = factor(case_when(
    sexnumber == 0 ~ "female",
    sexnumber == 1 ~ "male"
  ), levels = c("female", "male")))

rdata <- rdata %>% 
  rename_at('lopnr', ~'eid') %>% 
  rename_at('sex', ~'sexnumber')
vdata <- vdata %>% 
  select(-sexnumber) %>% 
  rename_at('sex', ~'sexnumber')

rdata_mm <- subset(rdata, rdata$mmcount_nockd >1)
rdata_nomm <- subset(rdata, rdata$mmcount_nockd<=1)
vdata_mm <- subset(vdata, vdata$mmcount_nockd >1)
vdata_nomm <- subset(vdata, vdata$mmcount_nockd<=1)
```



### Descriptive statistics

```{r, import}
rsel <- rdata[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]
vsel <- vdata[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]


rsel_mm <- rdata_mm[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]
vsel_mm <- vdata_mm[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]

rsel_nomm <- rdata_nomm[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]
vsel_nomm <- vdata_nomm[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2")]

rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind(rsel, vsel)
cdata$dt <- factor(
  cdata$dt,
  levels = c(1, 2),
  labels = c("Development data", "Validation data")
)
label(cdata$age) <- "Age"
label(cdata$sexnumber) <- "Sex"
label(cdata$egfrcr09_race_allnonblack) <- "eGFR creatinine"
label(cdata$acr2) <- "uACR"
units(cdata$age) <- "years"
units(cdata$egfrcr09_race_allnonblack) <- "ml/min/1.73m2"
units(cdata$acr2) <- "mg/mmol"
```


```{r tab1_bis}
gtsummary::tbl_summary(
  data = cdata |> select(-eid),
  label = list(age ~ "Age (years)", egfrcr09_race_allnonblack ~ "eGFR creatinine (ml/min/1.73m2)", acr2 ~"uACR (mg/mmol)"),
  by = "dt",
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
) |>
  gtsummary::as_kable_extra() |>
  kableExtra::kable_styling("striped")
```

## Goal 1 - develop a competing risks prediction model

### 1.1 Cumulative incidence curves

First, we draw the cumulative incidence curves of kidney failure.

<details>
  <summary>Click to expand code</summary>
```{r, cuminc, fig.align='center', eval=FALSE, echo=TRUE}

rdata <- rdata %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event", "acr2")

rdata <- rdata %>%
  mutate(time = if_else(time <=0, 0.1, time))

rdata <- as.data.frame(rdata)




# Expand data to prepare for fitting the model
rdata.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = rdata
)
# Save extended data with weights for KF (failcode=1)
# and mortality (failcode=2)
rdata.w1 <- rdata.w |> filter(failcode == 1)
rdata.w2 <- rdata.w |> filter(failcode == 2)

vdata <- vdata %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event") %>% 
  mutate(time)

vdata <- vdata %>%
  mutate(time = if_else(time <=0, 0.1, time))

vdata <- as.data.frame(vdata)

vdata.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = vdata
)
vdata.w1 <- vdata.w |> filter(failcode == 1)
vdata.w2 <- vdata.w |> filter(failcode == 2)

# Development set 

mfit_rdata <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = rdata.w1, 
  weights = weight.cens
)

mfit_vdata <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata.w1, 
  weights = weight.cens
)

par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
plot(
  mfit_rdata,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kdiney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event", 
  conf.int = TRUE
)
title("Development data")
plot(
  mfit_vdata,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kidney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("Validation data")
par(oldpar)

# Cumulative incidences for kidney failure
smfit_rdata <- summary(mfit_rdata, times = c(1, 2, 3, 4, 5))
smfit_vdata <- summary(mfit_vdata, times = c(1, 2, 3, 4, 5))

###Death

# Development set 
mfit_rdata2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = rdata.w2, 
  weights = weight.cens
)
mfit_vdata2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = vdata.w2, 
  weights = weight.cens
)
par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
plot(
  mfit_rdata2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event", 
  conf.int = TRUE
)
title("Development data")
plot(
  mfit_vdata2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("Validation data")
par(oldpar)

# Cumulative incidences for death
smfit_rdata2 <- summary(mfit_rdata2, times = c(1, 2, 3, 4, 5))
smfit_vdata2 <- summary(mfit_vdata2, times = c(1, 2, 3, 4, 5))
```

```{r, cuminc, fig.align='center', eval=FALSE, echo=TRUE}
#Multimorbidity vs no multimorbidity

rdata_mm <- rdata_mm %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event", "acr2")

rdata_mm <- rdata_mm %>%
  mutate(time = if_else(time <=0, 0.1, time))

rdata_mm <- as.data.frame(rdata_mm)

rdata_nomm <- rdata_nomm %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event", "acr2")

rdata_nomm <- rdata_nomm %>%
  mutate(time = if_else(time <=0, 0.1, time))

rdata_nomm <- as.data.frame(rdata_nomm)


# Expand data to prepare for fitting the model
rdata_mm.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = rdata_mm
)
rdata_nomm.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = rdata_nomm
)
# Save extended data with weights for KF (failcode=1)
# and mortality (failcode=2)
rdata_mm.w1 <- rdata_mm.w |> filter(failcode == 1)
rdata_mm.w2 <- rdata_mm.w |> filter(failcode == 2)

rdata_nomm.w1 <- rdata_nomm.w |> filter(failcode == 1)
rdata_nomm.w2 <- rdata_nomm.w |> filter(failcode == 2)

vdata_mm <- vdata_mm %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event") %>% 
  mutate(time)

vdata_mm <- vdata_mm %>%
  mutate(time = if_else(time <=0, 0.1, time))

vdata_mm <- as.data.frame(vdata_mm)

vdata_mm.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = vdata_mm
)
vdata_mm.w1 <- vdata_mm.w |> filter(failcode == 1)
vdata_mm.w2 <- vdata_mm.w |> filter(failcode == 2)

vdata_nomm <- vdata_nomm %>% 
  select("eid", "status", "time", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "event") %>% 
  mutate(time)

vdata_nomm <- vdata_nomm %>%
  mutate(time = if_else(time <=0, 0.1, time))

vdata_nomm <- as.data.frame(vdata_nomm)

vdata_nomm.w <- crprep(
  Tstop = "time",
  status = "event",
  trans = c(1, 2),
  id = "eid",
  keep = c("age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2"),
  data = vdata_nomm
)
vdata_nomm.w1 <- vdata_nomm.w |> filter(failcode == 1)
vdata_nomm.w2 <- vdata_nomm.w |> filter(failcode == 2)

# Development set 

mfit_rdata_mm <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = rdata_mm.w1, 
  weights = weight.cens
)
mfit_rdata_nomm <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = rdata_nomm.w1, 
  weights = weight.cens
)


mfit_vdata_mm <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata_mm.w1, 
  weights = weight.cens
)

mfit_vdata_nomm <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata_nomm.w1, 
  weights = weight.cens
)


par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))

plot(
  mfit_rdata_mm,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kidney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("SCREAM multimorbidity cohort")
plot(
  mfit_rdata_nomm,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kidney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("SCREAM no multimorbidity cohort")

plot(
  mfit_vdata_mm,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kidney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("UK Biobank multimorbidity cohort")
plot(
  mfit_vdata_nomm,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence kidney failure", bty = "n",
  ylim = c(0, 0.05), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("UK Biobank no multimorbidity cohort")
par(oldpar)

# Cumulative incidences for kidney failure
smfit_rdata_mm <- summary(mfit_rdata_mm, times = c(1, 2, 3, 4, 5))
smfit_vdata_mm <- summary(mfit_vdata_mm, times = c(1, 2, 3, 4, 5))
smfit_rdata_nomm <- summary(mfit_rdata_nomm, times = c(1, 2, 3, 4, 5))
smfit_vdata_nomm <- summary(mfit_vdata_nomm, times = c(1, 2, 3, 4, 5))

###Death

# Development set 
mfit_rdata_mm2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = rdata_mm.w2, 
  weights = weight.cens
)
mfit_vdata_mm2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = vdata_mm.w2, 
  weights = weight.cens
)

mfit_rdata_nomm2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = rdata_nomm.w2, 
  weights = weight.cens
)
mfit_vdata_nomm2 <- survfit(
  Surv(Tstart, Tstop, status == 2) ~ 1,
  data = vdata_nomm.w2, 
  weights = weight.cens
)

par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
plot(
  mfit_rdata_mm2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event", 
  conf.int = TRUE
)
title("SCREAM multimorbidity cohort")
plot(
  mfit_vdata_mm2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("UK Biobank multimorbidity cohort")
plot(
  mfit_rdata_nomm2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event", 
  conf.int = TRUE
)
title("SCREAM no multimorbidity cohort")
plot(
  mfit_vdata_nomm2,
  col = 1, lwd = 2,
  xlab = "Time (years)",
  ylab = "Cumulative incidence death", bty = "n",
  ylim = c(0, 0.50), 
  xlim = c(0, 5), 
  fun = "event",
  conf.int = TRUE
)
title("UK Biobank no multimorbidity cohort")
par(oldpar)

# Cumulative incidences for death
smfit_rdata_mm2 <- summary(mfit_rdata_mm2, times = c(1, 2, 3, 4, 5))
smfit_vdata_mm2 <- summary(mfit_vdata_mm2, times = c(1, 2, 3, 4, 5))
smfit_rdata_nomm2 <- summary(mfit_rdata_nomm2, times = c(1, 2, 3, 4, 5))
smfit_vdata_nomm2 <- summary(mfit_vdata_nomm2, times = c(1, 2, 3, 4, 5))
```

</details>

```{r, cuminc, fig.align='center', eval=TRUE}
```

The R packages and functions `cmprsk::cuminc()` and `mstate::Cuminc()`are good and easy alternatives to estimate the cumulative incidence function. 

```{r, res_ci, fig.align='center'}
res_ci <- cbind(
  1 - smfit_rdata$surv,
  1 - smfit_rdata$upper,
  1 - smfit_rdata$lower,
  1 - smfit_vdata$surv,
  1 - smfit_vdata$upper,
  1 - smfit_vdata$lower
)

res_ci <- round(res_ci, 4)
rownames(res_ci) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "Development data" = 3, "Validation data" = 3))|>
  add_header_above(c("Cumulative incidence of kidney failure" = 7))

res_ci2 <- cbind(
  1 - smfit_rdata2$surv,
  1 - smfit_rdata2$upper,
  1 - smfit_rdata2$lower,
  1 - smfit_vdata2$surv,
  1 - smfit_vdata2$upper,
  1 - smfit_vdata2$lower
)
res_ci2 <- round(res_ci2, 4)
rownames(res_ci2) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci2) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci2, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "Development data" = 3, "Validation data" = 3))|>
  add_header_above(c("Cumulative incidence of death" = 7))
```

The 5-year cumulative incidence of kidney failure was 3.13% (95% CI: 2.95-3.32%) in the SCREAM cohort, and 1.27% (95% CI: 1.13-1.41%) in the UK Biobank cohort.
The 5-year cumulative incidence of death was 27.01% (95% CI: 26.53-27.48%) in the SCREAM cohort, and 5.67% (95% CI: 5.38-5.96%) in the UK Biobank cohort.

```{r, res_ci, fig.align='center'}
##Multimorbidity cohorts
res_ci <- cbind(
  1 - smfit_rdata_mm$surv,
  1 - smfit_rdata_mm$upper,
  1 - smfit_rdata_mm$lower,
  1 - smfit_vdata_mm$surv,
  1 - smfit_vdata_mm$upper,
  1 - smfit_vdata_mm$lower
)
res_ci <- round(res_ci, 4)
rownames(res_ci) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "SCREAM multimorbidity cohort" = 3, "UK Biobank multimorbidity cohort" = 3))|>
  add_header_above(c("Cumulative incidence of kidney failure" = 7))

res_ci2 <- cbind(
  1 - smfit_rdata_mm2$surv,
  1 - smfit_rdata_mm2$upper,
  1 - smfit_rdata_mm2$lower,
  1 - smfit_vdata_mm2$surv,
  1 - smfit_vdata_mm2$upper,
  1 - smfit_vdata_mm2$lower
)
res_ci2 <- round(res_ci2, 4)
rownames(res_ci2) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci2) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci2, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "SCREAM multimorbidity cohort" = 3, "UK Biobank multimorbidity cohort" = 3))|>
  add_header_above(c("Cumulative incidence of death" = 7))
```
The 5-year cumulative incidence of kidney failure was 2.95% (95% CI: 2.74-3.16%) in the SCREAM multimorbidity cohort, and 1.22% (95% CI: 1.06-1.39%) in the UK Biobank multimorbidity cohort.
The 5-year cumulative incidence of death was 34.0% (95% CI: 33.39-34.60%) in the SCREAM multimorbidity cohort, and 5.61% (95% CI: 5.27-5.95%) in the UK Biobank multimorbidity cohort.

```{r, res_ci, fig.align='center'}
##Multimorbidity cohorts
res_ci <- cbind(
  1 - smfit_rdata_nomm$surv,
  1 - smfit_rdata_nomm$upper,
  1 - smfit_rdata_nomm$lower,
  1 - smfit_vdata_nomm$surv,
  1 - smfit_vdata_nomm$upper,
  1 - smfit_vdata_nomm$lower
)
res_ci <- round(res_ci, 4)
rownames(res_ci) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "SCREAM no multimorbidity cohort" = 3, "UK Biobank no multimorbidity cohort" = 3))|>
  add_header_above(c("Cumulative incidence of kidney failure" = 7))

res_ci2 <- cbind(
  1 - smfit_rdata_nomm2$surv,
  1 - smfit_rdata_nomm2$upper,
  1 - smfit_rdata_nomm2$lower,
  1 - smfit_vdata_nomm2$surv,
  1 - smfit_vdata_nomm2$upper,
  1 - smfit_vdata_nomm2$lower
)
res_ci2 <- round(res_ci2, 4)
rownames(res_ci2) <- c("1-year", "2-year", "3-year", "4-year", "5-year")
colnames(res_ci2) <- rep(c("Estimate", "Lower .95", "Upper .95"), 2)
kable(res_ci2, row.names = TRUE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, "SCREAM no multimorbidity cohort" = 3, "UK Biobank no multimorbidity cohort" = 3))|>
  add_header_above(c("Cumulative incidence of death" = 7))
```
The 5-year cumulative incidence of kidney failure was 3.56% (95% CI: 3.20-3.91%) in the SCREAM no multimorbidity cohort, and 0.74% (95% CI: 0.57-0.91%) in the UK Biobank no multimorbidity cohort.
The 5-year cumulative incidence of death was 11.11% (95% CI: 10.50-11.71%) in the SCREAM no multimorbidity cohort, and 3.34% (95% CI: 2.98-3.70%) in the UK Biobank no multimorbidity cohort.

### 1.2 Check non-linearity of continuous predictors

Here we investigate the potential non-linear relation between continuous predictors (i.e. age, eGFR, uACR) and the outcomes. We apply three-knot restricted cubic splines using `rms::rcs()` function (details are given in e.g. Frank Harrell's book 'Regression Model Strategies (second edition)', page 27. We assess the potential non-linearity graphically (plotting the two continuous predictors against the log relative hazards (XB or linear predictor) of both event types. Also, we compare the models with and without splines based on the AIC using `stats::AIC()`.

<details>
  <summary>Click to expand code</summary>
```{r,ff, warning=FALSE, fig.align='center', eval=FALSE, echo=TRUE}
# Models without splines
fit_csh <- CSC(
  Hist(time, event) ~ age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
  data = rdata,
  fitter = "cph"
)

fit_csc1 <- fit_csh$models$`Cause 1`
fit_csc2 <- fit_csh$models$`Cause 2`

# Models with splines
dd <- datadist(rdata)
options(datadist = "dd")

# Kidney failure
fit_csc1_rcs <- cph(
  Surv(time, event == 1) ~ rcs(age, 3) + rcs(egfrcr09_race_allnonblack, 3) + rcs(lnacr2, 3) + sexnumber,
  x = T,
  y = T,
  surv = T,
  data = rdata
)
print(fit_csc1_rcs)
print(summary(fit_csc1_rcs))
print(anova(fit_csc1_rcs))
P_csc1_age_rcs <- Predict(fit_csc1_rcs, "age")
P_csc1_egfrcr_rcs <- Predict(fit_csc1_rcs, "egfrcr09_race_allnonblack")
P_csc1_acr_rcs <- Predict(fit_csc1_rcs, "lnacr2")
options(datadist = NULL)
# Mortality
dd <- datadist(rdata)
options(datadist = "dd")
fit_csc2_rcs <- cph(
  Surv(time, event == 2) ~ rcs(age, 3) + rcs(egfrcr09_race_allnonblack, 3) + rcs(lnacr2, 3) + sexnumber,
  x = T,
  y = T,
  surv = T,
  data = rdata
)
 print(fit_csc2_rcs)
 print(summary(fit_csc2_rcs))
 print(anova(fit_csc2_rcs))
P_csc2_age_rcs <- Predict(fit_csc2_rcs, "age")
P_csc2_egfrcr_rcs <- Predict(fit_csc2_rcs, "egfrcr09_race_allnonblack")
P_csc2_acr_rcs <- Predict(fit_csc1_rcs, "lnacr2")
options(datadist = NULL)
oldpar <- par(mfrow = c(2, 2), mar = c(5, 5, 1, 1))
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  P_csc1_age_rcs$age,
  P_csc1_age_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Age years",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(18, 99)
)
polygon(
  c(P_csc1_age_rcs$age, rev(P_csc1_age_rcs$age)),
  c(P_csc1_age_rcs$lower, rev(P_csc1_age_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc1_age_rcs$age,
  P_csc1_age_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Age years",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(18, 99)
)
title("Kidney failure")
# CSC 1- egfr
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  P_csc1_egfrcr_rcs$egfrcr09_race_allnonblack,
  P_csc1_egfrcr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "eGFR creatinine",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(10, 60)
)
polygon(
  c(P_csc1_egfrcr_rcs$egfrcr09_race_allnonblack, rev(P_csc1_egfrcr_rcs$egfrcr09_race_allnonblack)),
  c(P_csc1_egfrcr_rcs$lower, rev(P_csc1_egfrcr_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc1_egfrcr_rcs$egfrcr09_race_allnonblack,
  P_csc1_egfrcr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "eGFR creatinine",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(10, 60)
)
title("Kidney failure")
par(xaxs = "i", yaxs = "i", las = 1)
options(datadist = NULL)
# CSC 2- age
plot(
  P_csc2_age_rcs$age,
  P_csc2_age_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Age years",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(18, 99)
)
polygon(
  c(P_csc2_age_rcs$age, rev(P_csc2_age_rcs$age)),
  c(P_csc2_age_rcs$lower, rev(P_csc2_age_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc2_age_rcs$age,
  P_csc2_age_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "Age years",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(18, 99)
)
title("Mortality")
# CSC 2 - egfrcr09_race_allnonblack
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  P_csc2_egfrcr_rcs$egfrcr09_race_allnonblack,
  P_csc2_egfrcr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "eGFR creatinine",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(10, 60)
)
polygon(
  c(P_csc2_egfrcr_rcs$egfrcr09_race_allnonblack, rev(P_csc2_egfrcr_rcs$egfrcr09_race_allnonblack)),
  c(P_csc2_egfrcr_rcs$lower, rev(P_csc2_egfrcr_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc2_egfrcr_rcs$egfrcr09_race_allnonblack,
  P_csc2_egfrcr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "eGFR creatinine",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(10, 60)
)
title("Mortality")
options(datadist = NULL)
par(oldpar)

# CSC 2- acr2
plot(
  P_csc1_acr_rcs$lnacr2,
  P_csc1_acr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "uACR mg/mmol",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(-12, 7)
)
polygon(
  c(P_csc1_acr_rcs$lnacr2, rev(P_csc1_acr_rcs$lnacr2)),
  c(P_csc1_acr_rcs$lower, rev(P_csc1_acr_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc1_acr_rcs$lnacr2,
  P_csc1_acr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "uACR mg/mmol",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(-12, 7)
)
title("Kidney failure")
# CSC 2 - acr2
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  P_csc2_acr_rcs$lnacr2,
  P_csc2_acr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "uACR mg/mmol",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(-12, 7)
)
polygon(
  c(P_csc2_acr_rcs$lnacr2, rev(P_csc2_acr_rcs$lnacr2)),
  c(P_csc2_acr_rcs$lower, rev(P_csc2_acr_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(
  P_csc2_acr_rcs$lnacr2,
  P_csc2_acr_rcs$yhat,
  type = "l",
  lwd = 2,
  col = "blue",
  bty = "n",
  xlab = "uACR mg/mmol",
  ylab = "log Relative Hazard",
  ylim = c(-3, 3),
  xlim = c(-12, 7)
)
title("Mortality")
options(datadist = NULL)
par(oldpar)
```
</details>

```{r,ff, warning=FALSE, fig.align='center', eval=TRUE}
```

```{r, res_aic, fig.align='center'}
res_AIC <- matrix(c(
  AIC(fit_csc1), AIC(fit_csc1_rcs),
  AIC(fit_csc2), AIC(fit_csc2_rcs)
),
byrow = T,
ncol = 2,
nrow = 2,
dimnames =
  list(
    c(
      "Kidney failure specific hazard",
      "Mortality"
    ),
    c(
      "AIC without splines",
      "AIC with splines"
    )
  )
)
kable(res_AIC, row.names = TRUE) |>
  kable_styling("striped", position = "center")
```

### 1.3 Checking proportional hazards assumption

We now examine the fits further by checking the proportionality of the cause-specific hazards.

<details>
  <summary>Click to expand code</summary>
```{r,ph_csc1, message=FALSE, warning=FALSE,fig.align='center', echo=TRUE, eval=FALSE}
zp_csc1 <- cox.zph(fit_csc1, transform = "identity")
par(las = 1, xaxs = "i", yaxs = "i")
# c(bottom, left, top, right)
oldpar <- par(mfrow = c(2, 2), mar = c(5, 6.1, 3.1, 1))
sub_title <- c("Age", "Sex", "eGFR creatinine", "uACR")
for (i in 1:4) {
  plot(
    zp_csc1[i],
    resid = F,
    bty = "n",
    xlim = c(0, 5)
  )
  abline(0, 0, lty = 3)
  title(sub_title[i])
}
mtext("Kidney Failure", side = 3, line = -1, outer = TRUE, font = 2)
par(oldpar)
kable(round(zp_csc1$table, 3)) |>
  kable_styling("striped", position = "center")
```
</details>

```{r,ph_csc1, message=FALSE, warning=FALSE,fig.align='center', echo=FALSE, eval=TRUE}
```

<details>
  <summary>Click to expand code</summary>
```{r,ph_csc2, message=FALSE, warning=FALSE,fig.align='center',echo=TRUE,eval=FALSE}
zp_csc2 <- cox.zph(fit_csc2, transform = "identity")
par(las = 1, xaxs = "i", yaxs = "i")
# c(bottom, left, top, right)
oldpar <- par(mfrow = c(2, 2), mar = c(5, 6.1, 3.1, 1))
sub_title <- c("Age", "Sex", "eGFR creatinine", "uACR")
for (i in 1:4) {
  plot(
    zp_csc2[i],
    resid = F,
    bty = "n",
    xlim = c(0, 5)
  )
  abline(0, 0, lty = 3)
  title(sub_title[i])
}
mtext(
  "Mortality",
  side = 3,
  line = -1,
  outer = TRUE,
  font = 2
)
par(oldpar)
kable(round(zp_csc2$table, 3)) |>
  kable_styling("striped", position = "center")
```
</details>

```{r,ph_csc2, message=FALSE, warning=FALSE,fig.align='center',echo=FALSE,eval=TRUE}
```

### 1.4 Examine the fit of the models

+ Cox proportional hazard model for kidney failure

```{r, summary_csc1, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_csc1_cph <- cph(
  Surv(time, event == 1) ~ age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
  x = T,
  y = T,
  surv = T,
  data = rdata
)
print(fit_csc1_cph)
# print(summary(fit_csc1_cph))
options(datadist = NULL)
```

+ Cox proportional hazard model for mortality

```{r, summary_csc2, fig.align='center',warning=FALSE}
dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")
fit_csc2_cph <- cph(
  Surv(time, event == 2) ~ age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
  x = T,
  y = T,
  surv = T,
  data = rdata
)
print(fit_csc2_cph)
# print(summary(fit_csc2_cph))
options(datadist = NULL)
```
### 1.5 Plot of predictors vs estimated risk at 5 years in the validation data

<details>
  <summary>Click to expand code</summary>
```{r, plot_risk, fig.align='center',warning=FALSE, echo=TRUE, eval=FALSE}
# Models -------------
fit_csh <- CSC(
  formula = Hist(time, event) ~ age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
  data = rdata
)

# External validation at 5 years
horizon <- 5

# Calculate predicted probabilities
vdata$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = vdata,
  times = horizon
)

###SCREAM mm vs nomm
rdata_mm$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = rdata_mm,
  times = horizon
)

rdata_nomm$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = rdata_nomm,
  times = horizon
)

###UKB mm vs nomm

vdata_mm$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = vdata_mm,
  times = horizon
)

vdata_nomm$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = vdata_nomm,
  times = horizon
)


# Age
oldpar <- par(mfrow = c(2, 2))
par(xaxs = "i", yaxs = "i", las = 1)
plot(vdata$age,
  vdata$pred,
  bty = "n",
  xlim = c(18, 99),
  ylim = c(0, 1),
  xlab = "Age, years",
  ylab = "Estimated risk"
)
lines(
  lowess(vdata$age, vdata$pred),
  col = "red",
  lwd = 2
)

# eGFR
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  vdata$egfrcr09_race_allnonblack,
  vdata$pred,
  bty = "n",
  xlim = c(10, 60),
  ylim = c(0, 1),
  xlab = "eGFR creatinine",
  ylab = "Estimated risk"
)
lines(
  lowess(vdata$egfrcr09_race_allnonblack, vdata$pred),
  col = "red",
  lwd = 2
)

# uACR
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  vdata$lnacr2,
  vdata$pred,
  xlim = c(0, 2150),
  ylim = c(0, 1),
  bty = "n",
  xlab = "uACR mg/mmol",
  ylab = "Estimated risk"
)
lines(
  lowess(vdata$lnacr2, vdata$pred),
  col = "red",
  lwd = 2
)

# Sex
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  vdata$sexnumber,
  vdata$pred,
  ylim = c(0, 1),
  bty = "n",
  xlab = "Sex (male)",
  ylab = "Estimated risk"
)
par(oldpar)
```
</details>

```{r, plot_risk, fig.align='center',warning=FALSE, echo=FALSE, eval=TRUE}
```

## Goal 2 - Assessing performance of a competing risks prediction model

Here we evaluate the performance of the prediction model in terms of calibration, discrimination and overall prediction error. 


### 2.1 Calibration

We assess calibration by:

- The calibration plot as a graphical representation of calibration using
 the pseudo observations and the subdistribution hazard approach
 - Numerical summaries of calibration:  
 - The observed vs expected ratio (O/E ratio) ;  
 - The squared bias, i.e., the average squared difference between primary event indicators and risk predictions;
 - The integrated Calibration Index (ICI), i.e., the average absolute difference between primary event indicators and risk predictions;  
 - E50, E90 and Emax denote the median, 90th percentile and the maximum of the absolute differences between primary event indicators and risk predictions;  
 - Calibration intercept/slope estimated using pseudo observations:  
 - If on average the risk estimates equal the primary event indicators, the calibration intercept will be zero. A negative calibration intercept indicates that the risk estimates are on average too high and a positive intercept indicates that the risk estimates are on average too low.   
 - A calibration slope between 0 and 1 indicates overfitting of the model, i.e., too extreme predictions, both on the low and on the high         end. A calibration slope >1 indicates predictions do not show enough variation.  
      
#### 2.1.1 Calibration using pseudo observations

We calculate calibration plot and numerical summaries of calibration as
ICI, E50, E90, Emax and root squared bias using pseudo value approach.

##### 2.1.1.1 Calibration plot using pseudo observations

<details>
  <summary>Click to expand code</summary>
```{r, cal, fig.align='center',warning=FALSE,message=FALSE,eval=FALSE,echo=TRUE, fig.width=10, fig.height=6}
# Models ----------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)


# Predicted risk estimation
pred <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata
)
pred_SCREAM_mm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = rdata_mm
)
pred_SCREAM_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = rdata_nomm
)
pred_UKB_mm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata_mm
)
pred_UKB_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata_nomm
)


# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
non_na_indices <- which(!is.na(vdata$pred))
vdata <- vdata %>%
  filter(!is.nan(pred))
non_na_indices <- which(!is.na(vdata$pred))
vdata <- vdata %>%
  filter(!is.nan(pred))

score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

calplot_pseudo <- plotCalibration(
  x = score_vdata,
  brier.in.legend = FALSE,
  auc.in.legend = FALSE,
  cens.method = "pseudo",
  bandwidth = 0.05, # leave as NULL for default choice of smoothing
  cex = 1,
  round = FALSE, # Important, keeps all unique risk estimates rather than rounding
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  rug = TRUE,
  xlab = "Predictions",
  bty = "n"
)
title("Calibration plot using pseudo observations")

score_rdata_mm <- Score(
  list("csh_validation_mm" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_mm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

calplot_pseudo_SCREAM_mm <- plotCalibration(
  x = score_rdata_mm,
  brier.in.legend = FALSE,
  auc.in.legend = FALSE,
  cens.method = "pseudo",
  bandwidth = 0.05, # leave as NULL for default choice of smoothing
  cex = 1,
  round = FALSE, # Important, keeps all unique risk estimates rather than rounding
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  rug = TRUE,
  xlab = "Predictions",
  bty = "n"
)
title("Calibration plot using pseudo observations - SCREAM multimorbidity cohort")

score_rdata_nomm <- Score(
  list("csh_validation_nomm" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_nomm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

calplot_pseudo_SCREAM_nomm <- plotCalibration(
  x = score_rdata_nomm,
  brier.in.legend = FALSE,
  auc.in.legend = FALSE,
  cens.method = "pseudo",
  bandwidth = 0.05, # leave as NULL for default choice of smoothing
  cex = 1,
  round = FALSE, # Important, keeps all unique risk estimates rather than rounding
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  rug = TRUE,
  xlab = "Predictions",
  bty = "n"
)
title("Calibration plot using pseudo observations - SCREAM no multimorbidity cohort")




score_vdata_mm <- Score(
  list("csh_validation_UKB_mm" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_mm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

calplot_pseudo_UKB_mm <- plotCalibration(
  x = score_vdata_mm,
  brier.in.legend = FALSE,
  auc.in.legend = FALSE,
  cens.method = "pseudo",
  bandwidth = 0.05, # leave as NULL for default choice of smoothing
  cex = 1,
  round = FALSE, # Important, keeps all unique risk estimates rather than rounding
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  rug = TRUE,
  xlab = "Predictions",
  bty = "n"
)
title("Calibration plot using pseudo observations - UK Biobank multimorbidity cohort")

score_vdata_nomm <- Score(
  list("csh_validation_UKB_nomm" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_nomm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

calplot_pseudo_UKB_nomm <- plotCalibration(
  x = score_vdata_nomm,
  brier.in.legend = FALSE,
  auc.in.legend = FALSE,
  cens.method = "pseudo",
  bandwidth = 0.05, # leave as NULL for default choice of smoothing
  cex = 1,
  round = FALSE, # Important, keeps all unique risk estimates rather than rounding
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  rug = TRUE,
  xlab = "Predictions",
  bty = "n"
)
title("Calibration plot using pseudo observations - UK Biobank no multimorbidity cohort")
```
</details>

```{r, cal, fig.align='center',warning=FALSE,message=FALSE,eval=TRUE,echo=FALSE, fig.width=15, fig.height=6}
```


##### 2.1.1.2 Numerical summaries of calibration using pseudo observations

<details>
  <summary>Click to expand code</summary>
```{r, num_cal_pseudo, fig.align='center',warning=FALSE,message=FALSE,echo=TRUE,eval=TRUE}

# We can extract predicted and observed, observed will depend on degree of smoothing (bandwidth)
dat_pseudo <- calplot_pseudo$plotFrames$csh_validation
dat_pseudo_SCREAM_mm <-calplot_pseudo_SCREAM_mm$plotFrames$csh_validation_mm
dat_pseudo_SCREAM_nomm <-calplot_pseudo_SCREAM_nomm$plotFrames$csh_validation_nomm
dat_pseudo_UKB_mm <-calplot_pseudo_UKB_mm$plotFrames$csh_validation_UKB_mm
dat_pseudo_UKB_nomm <-calplot_pseudo_UKB_nomm$plotFrames$csh_validation_UKB_nomm
#

######Does not work if you don't remove missing values earlier in code
# Calculate difference between predicted and observed (make sure to use all estimated risks, not just unique ones)
diff_pseudo <- pred - dat_pseudo$Obs[match(pred, dat_pseudo$Pred)]
diff_pseudo_SCREAM_mm <- pred_SCREAM_mm - dat_pseudo_SCREAM_mm$Obs[match(pred_SCREAM_mm, dat_pseudo_SCREAM_mm$Pred)]
diff_pseudo_SCREAM_nomm <- pred_SCREAM_nomm - dat_pseudo_SCREAM_nomm$Obs[match(pred_SCREAM_nomm, dat_pseudo_SCREAM_nomm$Pred)]
diff_pseudo_UKB_mm <- pred_UKB_mm - dat_pseudo_UKB_mm$Obs[match(pred_UKB_mm, dat_pseudo_UKB_mm$Pred)]
diff_pseudo_UKB_nomm <- pred_UKB_nomm - dat_pseudo_UKB_nomm$Obs[match(pred_UKB_nomm, dat_pseudo_UKB_nomm$Pred)]


# Collect all numerical summary measures

numsum_pseudo <- c(
  "ICI" = mean(abs(diff_pseudo)),
  setNames(quantile(abs(diff_pseudo), c(0.5, 0.9)), c("E50", "E90")),
  "Emax" = max(abs(diff_pseudo)),
  "Root squared bias" = sqrt(mean(diff_pseudo^2))
)


numsum_pseudo <- data.frame(
  Measure = c("ICI", "E50", "E90", "Emax", "Root squared bias"),
  Pseudo = c(mean(abs(diff_pseudo)),
             quantile(abs(diff_pseudo), c(0.5)),
             quantile(abs(diff_pseudo), c(0.9)),
             max(abs(diff_pseudo)),
             sqrt(mean(diff_pseudo^2))),
  SCREAM_mm = c(mean(abs(diff_pseudo_SCREAM_mm)),
                quantile(abs(diff_pseudo_SCREAM_mm), c(0.5)),
                quantile(abs(diff_pseudo_SCREAM_mm), c(0.9)),
                max(abs(diff_pseudo_SCREAM_mm)),
                sqrt(mean(diff_pseudo_SCREAM_mm^2))),
  SCREAM_nomm = c(mean(abs(diff_pseudo_SCREAM_nomm)),
                  quantile(abs(diff_pseudo_SCREAM_nomm), c(0.5)),
                  quantile(abs(diff_pseudo_SCREAM_nomm), c(0.9)),
                  max(abs(diff_pseudo_SCREAM_nomm)),
                  sqrt(mean(diff_pseudo_SCREAM_nomm^2))),
  UKB_mm = c(mean(abs(diff_pseudo_UKB_mm)),
             quantile(abs(diff_pseudo_UKB_mm), c(0.5)),
             quantile(abs(diff_pseudo_UKB_mm), c(0.9)),
             max(abs(diff_pseudo_UKB_mm)),
             sqrt(mean(diff_pseudo_UKB_mm^2))),
  UKB_nomm = c(mean(abs(diff_pseudo_UKB_nomm)),
               quantile(abs(diff_pseudo_UKB_nomm), c(0.5)),
               quantile(abs(diff_pseudo_UKB_nomm), c(0.9)),
               max(abs(diff_pseudo_UKB_nomm)),
               sqrt(mean(diff_pseudo_UKB_nomm^2)))
)
```
</details>


```{r, res_cal_pseudo, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE}
# rootmeansquared, E50, E90, Emax, int, slope,
k <- 4
res_num <- rbind(
  "Calibration measures - pseudo observations" = numsum_pseudo
)


kable(res_num) |>
  kable_styling("striped", position = "center")
```

#### 2.1.2 Calibration using the subdistribution hazard approach
Here we assess calibration using the subdistribution hazard approach (Austin et al.)

##### 2.1.2.1 Calibration plot using the subdistribution hazard approach

<details>
  <summary>Click to expand code</summary>
```{r, plot_sd, fig.align='center',warning=FALSE,message=FALSE,eval=FALSE,echo=TRUE, fig.width=10, fig.height=5}
# Models ----------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)


# Calibration plot (flexible regression approach) -------------------------

# Add estimated risk and complementary log-log of it to dataset
vdata$pred <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata,
  times = horizon
)

vdata$cll_pred <- log(-log(1 - pred))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)

colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))

vdata_bis <- cbind.data.frame(vdata, rcs_vdata)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time, event)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred
)

# Calibration plot
dat_fgr <- dat_fgr[order(dat_fgr$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr$pred,
  y = dat_fgr$obs,
  type = "l",
  xlim = c(0, 0.6),
  ylim = c(0, 0.6),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach", 
      cex.main = .90)
```
</details>

```{r, plot_sd, fig.align='center',warning=FALSE,message=FALSE,eval=FALSE,echo=TRUE}
## MM & noMM cohorts

# Calibration plot (flexible regression approach) -------------------------

# Add estimated risk and complementary log-log of it to dataset
rdata_mm$pred_SCREAM_mm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = rdata_mm,
  times = horizon
)

rdata_mm$cll_pred <- log(-log(1 - pred_SCREAM_mm))

rdata_nomm$pred_SCREAM_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = rdata_nomm,
  times = horizon
)

rdata_nomm$cll_pred <- log(-log(1 - pred_SCREAM_nomm))


vdata_mm$pred_UKB_mm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata_mm,
  times = horizon
)

vdata_mm$cll_pred <- log(-log(1 - pred_UKB_mm))

vdata_nomm$pred_UKB_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata_nomm,
  times = horizon
)

vdata_nomm$cll_pred <- log(-log(1 - pred_UKB_nomm))

# 5 knots seems to give somewhat equivalent graph to pseudo method with bw = 0.05
n_internal_knots <- 5 # Austin et al. advise to use between 3 (more smoothing, less flexible) and 5 (less smoothing, more flexible)
rcs_vdata <- ns(vdata$cll_pred, df = n_internal_knots + 1)

colnames(rcs_vdata) <- paste0("basisf_", colnames(rcs_vdata))

vdata_bis <- cbind.data.frame(vdata, rcs_vdata)


rcs_rdata_mm <- ns(rdata_mm$cll_pred, df = n_internal_knots + 1)

colnames(rcs_rdata_mm) <- paste0("basisf_", colnames(rcs_rdata_mm))

rdata_bis_mm <- cbind.data.frame(rdata_mm, rcs_rdata_mm)

rcs_rdata_nomm <- ns(rdata_nomm$cll_pred, df = n_internal_knots + 1)

colnames(rcs_rdata_nomm) <- paste0("basisf_", colnames(rcs_rdata_nomm))

rdata_bis_nomm <- cbind.data.frame(rdata_nomm, rcs_rdata_nomm)


rcs_vdata_mm <- ns(vdata_mm$cll_pred, df = n_internal_knots + 1)

colnames(rcs_vdata_mm) <- paste0("basisf_", colnames(rcs_vdata_mm))

vdata_bis_mm <- cbind.data.frame(vdata_mm, rcs_vdata_mm)

rcs_vdata_nomm <- ns(vdata_nomm$cll_pred, df = n_internal_knots + 1)

colnames(rcs_vdata_nomm) <- paste0("basisf_", colnames(rcs_vdata_nomm))

vdata_bis_nomm <- cbind.data.frame(vdata_nomm, rcs_vdata_nomm)

# Use subdistribution hazards (Fine-Gray) model
form_fgr <- reformulate(
  termlabels = colnames(rcs_vdata),
  response = "Hist(time, event)"
)

# Regress subdistribution of event of interest on cloglog of estimated risks
calib_fgr <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis
)

calib_fgr_SCREAM_mm <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = rdata_bis_mm
)
calib_fgr_SCREAM_nomm <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = rdata_bis_nomm
)
calib_fgr_UKB_mm <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis_mm
)
calib_fgr_UKB_nomm <- FGR(
  formula = form_fgr,
  cause = primary_event,
  data = vdata_bis_nomm
)

# Add observed and predicted together in a data frame
dat_fgr <- cbind.data.frame(
  "obs" = predict(calib_fgr, times = horizon, newdata = vdata_bis),
  "pred" = vdata$pred)

dat_fgr_SCREAM_mm <- cbind.data.frame(
  "obs" = predict(calib_fgr_SCREAM_mm, times = horizon, newdata = rdata_bis_mm),
  "pred" = rdata_mm$pred) 

dat_fgr_SCREAM_nomm <- cbind.data.frame(
  "obs" = predict(calib_fgr_SCREAM_nomm, times = horizon, newdata = rdata_bis_nomm),
  "pred" = rdata_nomm$pred) 

dat_fgr_UKB_mm <- cbind.data.frame(
  "obs" = predict(calib_fgr_UKB_mm, times = horizon, newdata = vdata_bis_mm),
  "pred" = vdata_mm$pred) 

dat_fgr_UKB_nomm <- cbind.data.frame(
  "obs" = predict(calib_fgr_UKB_nomm, times = horizon, newdata = vdata_bis_nomm),
  "pred" = vdata_nomm$pred) 

# Calibration plot
dat_fgr <- dat_fgr[order(dat_fgr$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr$pred,
  y = dat_fgr$obs,
  type = "l",
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach", 
      cex.main = .90)


dat_fgr_SCREAM_mm <- dat_fgr_SCREAM_mm[order(dat_fgr_SCREAM_mm$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr_SCREAM_mm$pred,
  y = dat_fgr_SCREAM_mm$obs,
  type = "l",
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach:\nSCREAM multimorbidity cohort", 
      cex.main = .90)

dat_fgr_SCREAM_nomm <- dat_fgr_SCREAM_nomm[order(dat_fgr_SCREAM_nomm$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr_SCREAM_nomm$pred,
  y = dat_fgr_SCREAM_nomm$obs,
  type = "l",
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach:\nSCREAM no multimorbidity cohort", 
      cex.main = .90)

dat_fgr_UKB_mm <- dat_fgr_UKB_mm[order(dat_fgr_UKB_mm$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr_UKB_mm$pred,
  y = dat_fgr_UKB_mm$obs,
  type = "l",
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach:\nUK Biobank multimorbidity cohort", 
      cex.main = .90)

dat_fgr_UKB_nomm <- dat_fgr_UKB_nomm[order(dat_fgr_UKB_nomm$pred), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = dat_fgr_UKB_nomm$pred,
  y = dat_fgr_UKB_nomm$obs,
  type = "l",
  xlim = c(0, 1.0),
  ylim = c(0, 1.0),
  xlab = "Predictions",
  ylab = "Observed outcome proportion",
  bty = "n"
)
abline(a = 0, b = 1, lty = "dashed", col = "red")
title("Calibration plot using subdistribution hazard approach:\nUK Biobank no multimorbidity cohort", 
      cex.main = .90)
```

```{r, plot_sd, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE,eval=TRUE, fig.width=5, fig.height=5}
```


##### 2.1.2.2 Numerical summaries of calibration using the subdistribution hazard approach
Here we assess calibration using the subdistribution hazard approach (Austin et al.)

<details>
  <summary>Click to expand code</summary>
```{r, num_cal_sd, fig.align='center',warning=FALSE,message=FALSE,eval=FALSE, echo=TRUE}
# Numerical summary measures
diff_fgr <- dat_fgr$pred - dat_fgr$obs
diff_fgr_SCREAM_mm <- dat_fgr_SCREAM_mm$pred - dat_fgr_SCREAM_mm$obs
diff_fgr_SCREAM_nomm <- dat_fgr_SCREAM_nomm$pred - dat_fgr_SCREAM_nomm$obs
diff_fgr_UKB_mm <- dat_fgr_UKB_mm$pred - dat_fgr_UKB_mm$obs
diff_fgr_UKB_nomm <- dat_fgr_UKB_nomm$pred - dat_fgr_UKB_nomm$obs

numsum_fgr <- c(
  "ICI" = mean(abs(diff_fgr)),
  setNames(quantile(abs(diff_fgr), c(0.5, 0.9)), c("E50", "E90")),
  "Emax" = max(abs(diff_fgr)),
  "Root squared bias" = sqrt(mean(diff_fgr^2))
)


numsum_fgr <- data.frame(
  Measure = c("ICI", "E50", "E90", "Emax", "Root squared bias"),
  UKB_validation = c(mean(abs(diff_fgr)),
             quantile(abs(diff_fgr), c(0.5)),
             quantile(abs(diff_fgr), c(0.9)),
             max(abs(diff_fgr)),
             sqrt(mean(diff_fgr^2))),
  SCREAM_mm = c(mean(abs(diff_fgr_SCREAM_mm)),
                quantile(abs(diff_fgr_SCREAM_mm), c(0.5)),
                quantile(abs(diff_fgr_SCREAM_mm), c(0.9)),
                max(abs(diff_fgr_SCREAM_mm)),
                sqrt(mean(diff_fgr_SCREAM_mm^2))),
  SCREAM_nomm = c(mean(abs(diff_fgr_SCREAM_nomm)),
                  quantile(abs(diff_fgr_SCREAM_nomm), c(0.5)),
                  quantile(abs(diff_fgr_SCREAM_nomm), c(0.9)),
                  max(abs(diff_fgr_SCREAM_nomm)),
                  sqrt(mean(diff_fgr_SCREAM_nomm^2))),
  UKB_mm = c(mean(abs(diff_fgr_UKB_mm)),
             quantile(abs(diff_fgr_UKB_mm), c(0.5)),
             quantile(abs(diff_fgr_UKB_mm), c(0.9)),
             max(abs(diff_fgr_UKB_mm)),
             sqrt(mean(diff_fgr_UKB_mm^2))),
  UKB_nomm = c(mean(abs(diff_fgr_UKB_nomm)),
               quantile(abs(diff_fgr_UKB_nomm), c(0.5)),
               quantile(abs(diff_fgr_UKB_nomm), c(0.9)),
               max(abs(diff_fgr_UKB_nomm)),
               sqrt(mean(diff_fgr_UKB_nomm^2)))
)
```
</details>

```{r, num_cal_sd, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE,eval=TRUE, fig.width=5, fig.height=5}
```

```{r, num_cal_sd_out, fig.align='center',warning=FALSE,message=FALSE,echo=FALSE}
k <- 4
res_num <- rbind(
  "Calibration measures - subdistribution " = numsum_fgr
)

kable(res_num) |>
  kable_styling("striped", position = "center")



```
#### 2.1.2.3 Calibration plot using pseudo-observations (LOESS smoothing)

<details>
  <summary>Click to expand code</summary>
```{r loess_pseudo, echo=TRUE, eval=FALSE, fig.width=5, fig.height=5, fig.align='center'}
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos <- pseudos[order(pseudos$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos, degree = 1, span = 0.33), 
  se = TRUE
)

# Calibration plot (reported in manuscript):

# First, prepare histogram of estimated risks for x-axis
spike_bounds <- c(-0.075, 0)
bin_breaks <- seq(0, 0.6, length.out = 100 + 1)
freqs <- table(cut(pred, breaks = bin_breaks))
bins <- bin_breaks[-1]
freqs_valid <- freqs[freqs > 0]
freqs_rescaled <- spike_bounds[1] + (spike_bounds[2] - spike_bounds[1]) * 
  (freqs_valid - min(freqs_valid)) / (max(freqs_valid) - min(freqs_valid))

# Produce plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = pseudos$risk, 
  y = pseudos$pseudovalue,
  xlim = c(0, 0.6), 
  ylim = c(spike_bounds[1], 0.6),
  yaxt = "n",
  frame.plot = FALSE,
  xlab = "Estimated risks",
  ylab = "Observed outcome proportions", 
  type = "n"
)
axis(2, seq(0, 0.6, by = 0.1), labels = seq(0, 0.6, by = 0.1))
polygon(
  x = c(pseudos$risk, rev(pseudos$risk)),
  y = c(
    pmax(smooth_pseudos$fit - qt(0.975, smooth_pseudos$df) * smooth_pseudos$se, 0),
    rev(smooth_pseudos$fit + qt(0.975, smooth_pseudos$df) * smooth_pseudos$se)
  ),
  border = FALSE,
  col = "lightgray"
)
abline(a = 0, b = 1, col = "gray")
lines(x = pseudos$risk, y = smooth_pseudos$fit, lwd = 2)
segments(
  x0 = bins[freqs > 0], 
  y0 = spike_bounds[1], 
  x1 = bins[freqs > 0], 
  y1 = freqs_rescaled
)
title("Calibration plot using pseudo-observations (LOESS smoothing):\nvdata", cex.main = .90)

```
</details>

```{r loess_pseudo, echo=TRUE, eval=FALSE, fig.width=5, fig.height=5, fig.align='center'}
####SCREAM Multimorbidity
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos_SCREAM_mm <- data.frame(score_vdata_mm$Calibration$plotframe)
pseudos_SCREAM_mm <- pseudos_SCREAM_mm[order(pseudos_SCREAM_mm$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos_SCREAM_mm <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos_SCREAM_mm, degree = 1, span = 0.33), 
  se = TRUE
)

# Calibration plot (reported in manuscript):

# First, prepare histogram of estimated risks for x-axis
spike_bounds <- c(-0.075, 0)
bin_breaks <- seq(0, 0.6, length.out = 100 + 1)
freqs <- table(cut(pred, breaks = bin_breaks))
bins <- bin_breaks[-1]
freqs_valid <- freqs[freqs > 0]
freqs_rescaled <- spike_bounds[1] + (spike_bounds[2] - spike_bounds[1]) * 
  (freqs_valid - min(freqs_valid)) / (max(freqs_valid) - min(freqs_valid))

# Produce plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = pseudos_SCREAM_mm$risk, 
  y = pseudos_SCREAM_mm$pseudovalue,
  xlim = c(0, 0.6), 
  ylim = c(spike_bounds[1], 0.6),
  yaxt = "n",
  frame.plot = FALSE,
  xlab = "Estimated risks",
  ylab = "Observed outcome proportions", 
  type = "n"
)
axis(2, seq(0, 0.6, by = 0.1), labels = seq(0, 0.6, by = 0.1))
polygon(
  x = c(pseudos_SCREAM_mm$risk, rev(pseudos_SCREAM_mm$risk)),
  y = c(
    pmax(smooth_pseudos_SCREAM_mm$fit - qt(0.975, smooth_pseudos_SCREAM_mm$df) * smooth_pseudos_SCREAM_mm$se, 0),
    rev(smooth_pseudos_SCREAM_mm$fit + qt(0.975, smooth_pseudos_SCREAM_mm$df) * smooth_pseudos_SCREAM_mm$se)
  ),
  border = FALSE,
  col = "lightgray"
)
abline(a = 0, b = 1, col = "gray")
lines(x = pseudos_SCREAM_mm$risk, y = smooth_pseudos_SCREAM_mm$fit, lwd = 2)
segments(
  x0 = bins[freqs > 0], 
  y0 = spike_bounds[1], 
  x1 = bins[freqs > 0], 
  y1 = freqs_rescaled
)
title("Calibration plot using pseudo-observations (LOESS smoothing):\nSCREAM multimorbidity cohort", cex.main = .90)
```

```{r loess_pseudo, echo=TRUE, eval=FALSE, fig.width=5, fig.height=5, fig.align='center'}
####SCREAM NO Multimorbidity
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos_SCREAM_nomm <- data.frame(score_vdata_nomm$Calibration$plotframe)
pseudos_SCREAM_nomm <- pseudos_SCREAM_nomm[order(pseudos_SCREAM_nomm$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos_SCREAM_nomm <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos_SCREAM_nomm, degree = 1, span = 0.33), 
  se = TRUE
)

# Calibration plot (reported in manuscript):

# First, prepare histogram of estimated risks for x-axis
spike_bounds <- c(-0.075, 0)
bin_breaks <- seq(0, 0.6, length.out = 100 + 1)
freqs <- table(cut(pred, breaks = bin_breaks))
bins <- bin_breaks[-1]
freqs_valid <- freqs[freqs > 0]
freqs_rescaled <- spike_bounds[1] + (spike_bounds[2] - spike_bounds[1]) * 
  (freqs_valid - min(freqs_valid)) / (max(freqs_valid) - min(freqs_valid))

# Produce plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = pseudos_SCREAM_nomm$risk, 
  y = pseudos_SCREAM_nomm$pseudovalue,
  xlim = c(0, 0.6), 
  ylim = c(spike_bounds[1], 0.6),
  yaxt = "n",
  frame.plot = FALSE,
  xlab = "Estimated risks",
  ylab = "Observed outcome proportions", 
  type = "n"
)
axis(2, seq(0, 0.6, by = 0.1), labels = seq(0, 0.6, by = 0.1))
polygon(
  x = c(pseudos_SCREAM_nomm$risk, rev(pseudos_SCREAM_nomm$risk)),
  y = c(
    pmax(smooth_pseudos_SCREAM_nomm$fit - qt(0.975, smooth_pseudos_SCREAM_nomm$df) * smooth_pseudos_SCREAM_nomm$se, 0),
    rev(smooth_pseudos_SCREAM_nomm$fit + qt(0.975, smooth_pseudos_SCREAM_nomm$df) * smooth_pseudos_SCREAM_nomm$se)
  ),
  border = FALSE,
  col = "lightgray"
)
abline(a = 0, b = 1, col = "gray")
lines(x = pseudos_SCREAM_nomm$risk, y = smooth_pseudos_SCREAM_nomm$fit, lwd = 2)
segments(
  x0 = bins[freqs > 0], 
  y0 = spike_bounds[1], 
  x1 = bins[freqs > 0], 
  y1 = freqs_rescaled
)
title("Calibration plot using pseudo-observations (LOESS smoothing):\nSCREAM no multimorbidity cohort", cex.main = .90)
```

```{r loess_pseudo, echo=TRUE, eval=FALSE, fig.width=5, fig.height=5, fig.align='center'}
####UKB Multimorbidity
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos_UKB_mm <- data.frame(score_rdata_mm$Calibration$plotframe)
pseudos_UKB_mm <- pseudos_UKB_mm[order(pseudos_UKB_mm$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos_UKB_mm <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos_UKB_mm, degree = 1, span = 0.33), 
  se = TRUE
)

# Calibration plot (reported in manuscript):

# First, prepare histogram of estimated risks for x-axis
spike_bounds <- c(-0.075, 0)
bin_breaks <- seq(0, 0.6, length.out = 100 + 1)
freqs <- table(cut(pred, breaks = bin_breaks))
bins <- bin_breaks[-1]
freqs_valid <- freqs[freqs > 0]
freqs_rescaled <- spike_bounds[1] + (spike_bounds[2] - spike_bounds[1]) * 
  (freqs_valid - min(freqs_valid)) / (max(freqs_valid) - min(freqs_valid))

# Produce plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = pseudos_UKB_mm$risk, 
  y = pseudos_UKB_mm$pseudovalue,
  xlim = c(0, 0.6), 
  ylim = c(spike_bounds[1], 0.6),
  yaxt = "n",
  frame.plot = FALSE,
  xlab = "Estimated risks",
  ylab = "Observed outcome proportions", 
  type = "n"
)
axis(2, seq(0, 0.6, by = 0.1), labels = seq(0, 0.6, by = 0.1))
polygon(
  x = c(pseudos_UKB_mm$risk, rev(pseudos_UKB_mm$risk)),
  y = c(
    pmax(smooth_pseudos_UKB_mm$fit - qt(0.975, smooth_pseudos_UKB_mm$df) * smooth_pseudos_UKB_mm$se, 0),
    rev(smooth_pseudos_UKB_mm$fit + qt(0.975, smooth_pseudos_UKB_mm$df) * smooth_pseudos_UKB_mm$se)
  ),
  border = FALSE,
  col = "lightgray"
)
abline(a = 0, b = 1, col = "gray")
lines(x = pseudos_UKB_mm$risk, y = smooth_pseudos_UKB_mm$fit, lwd = 2)
segments(
  x0 = bins[freqs > 0], 
  y0 = spike_bounds[1], 
  x1 = bins[freqs > 0], 
  y1 = freqs_rescaled
)
title("Calibration plot using pseudo-observations (LOESS smoothing):\nUK Biobank multimorbidity cohort", cex.main = .90)
```

```{r loess_pseudo, echo=TRUE, eval=FALSE, fig.width=5, fig.height=5, fig.align='center'}
####UKB NO Multimorbidity
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos_UKB_nomm <- data.frame(score_rdata_nomm$Calibration$plotframe)
pseudos_UKB_nomm <- pseudos_UKB_nomm[order(pseudos_UKB_nomm$risk), ]

# Use linear loess (weighted local regression with polynomial degree = 1) smoothing
smooth_pseudos_UKB_nomm <- predict(
  stats::loess(pseudovalue ~ risk, data = pseudos_UKB_nomm, degree = 1, span = 0.33), 
  se = TRUE
)

# Calibration plot (reported in manuscript):

# First, prepare histogram of estimated risks for x-axis
spike_bounds <- c(-0.075, 0)
bin_breaks <- seq(0, 0.6, length.out = 100 + 1)
freqs <- table(cut(pred, breaks = bin_breaks))
bins <- bin_breaks[-1]
freqs_valid <- freqs[freqs > 0]
freqs_rescaled <- spike_bounds[1] + (spike_bounds[2] - spike_bounds[1]) * 
  (freqs_valid - min(freqs_valid)) / (max(freqs_valid) - min(freqs_valid))

# Produce plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  x = pseudos_UKB_nomm$risk, 
  y = pseudos_UKB_nomm$pseudovalue,
  xlim = c(0, 0.6), 
  ylim = c(spike_bounds[1], 0.6),
  yaxt = "n",
  frame.plot = FALSE,
  xlab = "Estimated risks",
  ylab = "Observed outcome proportions", 
  type = "n"
)
axis(2, seq(0, 0.6, by = 0.1), labels = seq(0, 0.6, by = 0.1))
polygon(
  x = c(pseudos_UKB_nomm$risk, rev(pseudos_UKB_nomm$risk)),
  y = c(
    pmax(smooth_pseudos_UKB_nomm$fit - qt(0.975, smooth_pseudos_UKB_nomm$df) * smooth_pseudos_UKB_nomm$se, 0),
    rev(smooth_pseudos_UKB_nomm$fit + qt(0.975, smooth_pseudos_UKB_nomm$df) * smooth_pseudos_UKB_nomm$se)
  ),
  border = FALSE,
  col = "lightgray"
)
abline(a = 0, b = 1, col = "gray")
lines(x = pseudos_UKB_nomm$risk, y = smooth_pseudos_UKB_nomm$fit, lwd = 2)
segments(
  x0 = bins[freqs > 0], 
  y0 = spike_bounds[1], 
  x1 = bins[freqs > 0], 
  y1 = freqs_rescaled
)
title("Calibration plot using pseudo-observations (LOESS smoothing):\nUK Biobank no multimorbidity cohort", cex.main = .90)
```

```{r loess_pseudo, echo=FALSE, eval=TRUE, fig.width=5, fig.height=5, fig.align='center'}
```


#### 2.1.3 Observed and Expected ratio

<details>
  <summary>Click to expand code</summary>
```{r, oe_ratio, fig.align='center',warning=FALSE,message=FALSE,echo=TRUE}

# Models ----------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Add estimated risk and complementary log-log of it to dataset
pred <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata,
  times = horizon
)

pred_SCREAM_mm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = rdata_mm,
  times = horizon
)
pred_SCREAM_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = rdata_nomm,
  times = horizon
)

pred_UKB_mm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata_mm,
  times = horizon
)
pred_UKB_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  newdata = vdata_nomm,
  times = horizon
)

## Observed/Expected ratio --------------------------------------------
# First calculate Aalen-Johansen estimate (as 'observed')
obj <- summary(survfit(Surv(time, status) ~ 1,
  data = vdata
),
times = horizon
)

aj <- list(
  "obs" = obj$pstate[, primary_event + 1],
  "se" = obj$std.err[, primary_event + 1]
)

obj_SCREAM_mm <- summary(survfit(Surv(time, status) ~ 1,
  data = rdata_mm
),
times = horizon
)

aj_SCREAM_mm <- list(
  "obs" = obj_SCREAM_mm$pstate[, primary_event + 1],
  "se" = obj_SCREAM_mm$std.err[, primary_event + 1]
)

obj_SCREAM_nomm <- summary(survfit(Surv(time, status) ~ 1,
  data = rdata_nomm
),
times = horizon
)

aj_SCREAM_nomm <- list(
  "obs" = obj_SCREAM_nomm$pstate[, primary_event + 1],
  "se" = obj_SCREAM_nomm$std.err[, primary_event + 1]
)


obj_UKB_mm <- summary(survfit(Surv(time, status) ~ 1,
  data = vdata_mm
),
times = horizon
)

aj_UKB_mm <- list(
  "obs" = obj_UKB_mm$pstate[, primary_event + 1],
  "se" = obj_UKB_mm$std.err[, primary_event + 1]
)

obj_UKB_nomm <- summary(survfit(Surv(time, status) ~ 1,
  data = vdata_nomm
),
times = horizon
)

aj_UKB_nomm <- list(
  "obs" = obj_UKB_nomm$pstate[, primary_event + 1],
  "se" = obj_UKB_nomm$std.err[, primary_event + 1]
)

# Calculate O/E
OE <- aj$obs / mean(pred)
OE_SCREAM_mm <- aj_SCREAM_mm$obs / mean(pred_SCREAM_mm)
OE_SCREAM_nomm <- aj_SCREAM_nomm$obs / mean(pred_SCREAM_nomm)
OE_UKB_mm <- aj_UKB_mm$obs / mean(pred_UKB_mm)
OE_UKB_nomm <- aj_UKB_nomm$obs / mean(pred_UKB_nomm)

# For the confidence interval we use method proposed in Debray et al. (2017) doi:10.1136/bmj.i6460
k <- 2
alpha <- 0.05
OE_summary <- data.frame(
  Measure = c("OE", "Lower .95", "Upper .95"),
  UKB_validation = c(OE,
             exp(log(OE) - qnorm(1 - alpha / 2) * aj$se / aj$obs),
             exp(log(OE) + qnorm(1 - alpha / 2) * aj$se / aj$obs)),
  SCREAM_mm = c(OE_SCREAM_mm,
             exp(log(OE_SCREAM_mm) - qnorm(1 - alpha / 2) * aj_SCREAM_mm$se / aj_SCREAM_mm$obs),
             exp(log(OE_SCREAM_mm) + qnorm(1 - alpha / 2) * aj_SCREAM_mm$se / aj_SCREAM_mm$obs)),
  SCREAM_nomm = c(OE_SCREAM_nomm,
             exp(log(OE_SCREAM_nomm) - qnorm(1 - alpha / 2) * aj_SCREAM_nomm$se / aj_SCREAM_nomm$obs),
             exp(log(OE_SCREAM_nomm) + qnorm(1 - alpha / 2) * aj_SCREAM_nomm$se / aj_SCREAM_nomm$obs)),
  UKB_mm = c(OE_UKB_mm,
             exp(log(OE_UKB_mm) - qnorm(1 - alpha / 2) * aj_UKB_mm$se / aj_UKB_mm$obs),
             exp(log(OE_UKB_mm) + qnorm(1 - alpha / 2) * aj_UKB_mm$se / aj_UKB_mm$obs)),
  UKB_nomm = c(OE_UKB_nomm,
             exp(log(OE_UKB_nomm) - qnorm(1 - alpha / 2) * aj_UKB_nomm$se / aj_UKB_nomm$obs),
             exp(log(OE_UKB_nomm) + qnorm(1 - alpha / 2) * aj_UKB_nomm$se / aj_UKB_nomm$obs))
)



```
</details>

```{r, res_OE, echo=FALSE}
# OE, Lower.95, Upper .95

kable(OE_summary) |>
  kable_styling("striped", position = "center")
```


#### 2.1.4 Calibration intercept and slope using pseudo observations


<details>
  <summary>Click to expand code</summary>
```{r, cal_intslope_pseudo, fig.align='center',warning=FALSE,message=FALSE, echo=TRUE}

# Models ----------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)


# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Predicted risk estimation
pred <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata
)

pred_SCREAM_mm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = rdata_mm
)

pred_SCREAM_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = rdata_nomm
)

pred_UKB_mm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata_mm
)

pred_UKB_nomm <- predictRisk(fit_csh,
  cause = primary_event,
  times = horizon,
  newdata = vdata_nomm
)
# Calibration plot (pseudo-obs approach) ----------------------------------
# First compute riskRegression::Score()
score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

score_rdata_SCREAM_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_mm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

score_rdata_SCREAM_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_nomm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

score_vdata_UKB_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_mm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

score_vdata_UKB_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_nomm,
  conf.int = TRUE,
  times = horizon,
  #  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)


## Calibration intercept and slope --------------------------------------
# Use pseudo-observations calculated by Score() (can alternatively use pseudo::pseudoci)
pseudos <- data.frame(score_vdata$Calibration$plotframe)
pseudos_SCREAM_mm <- data.frame(score_rdata_SCREAM_mm$Calibration$plotframe)
pseudos_SCREAM_nomm <- data.frame(score_rdata_SCREAM_nomm$Calibration$plotframe)
pseudos_UKB_mm <- data.frame(score_vdata_UKB_mm$Calibration$plotframe)
pseudos_UKB_nomm <- data.frame(score_vdata_UKB_nomm$Calibration$plotframe)

# Note:
# - 'pseudos' is the data.frame with ACTUAL pseudo-observations, not the smoothed ones
# - Column ID is not the id in vdata; it is just a number assigned to each row of
# the original validation data sorted by time and event indicator
pseudos$cll_pred <- log(-log(1 - pseudos$risk)) # add the cloglog risk ests
pseudos_SCREAM_mm$cll_pred <- log(-log(1 - pseudos_SCREAM_mm$risk))
pseudos_SCREAM_nomm$cll_pred <- log(-log(1 - pseudos_SCREAM_nomm$risk))
pseudos_UKB_mm$cll_pred <- log(-log(1 - pseudos_UKB_mm$risk))
pseudos_UKB_nomm$cll_pred <- log(-log(1 - pseudos_UKB_nomm$risk))
# Fit model for calibration intercept
fit_cal_int <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_int_SCREAM_mm <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos_SCREAM_mm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_int_SCREAM_nomm <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos_SCREAM_nomm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_int_UKB_mm <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos_UKB_mm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_int_UKB_nomm <- geese(
  pseudovalue ~ offset(cll_pred),
  data = pseudos_UKB_nomm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Fit model for calibration slope
fit_cal_slope <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_slope_SCREAM_mm <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos_SCREAM_mm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_slope_SCREAM_nomm <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos_SCREAM_nomm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_slope_UKB_mm <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos_UKB_mm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

fit_cal_slope_UKB_nomm <- geese(
  pseudovalue ~ offset(cll_pred) + cll_pred,
  data = pseudos_UKB_nomm,
  id = ID,
  scale.fix = TRUE,
  family = gaussian,
  mean.link = "cloglog",
  corstr = "independence",
  jack = TRUE
)

# Perform joint test on intercept and slope
betas <- fit_cal_slope$beta
vcov_mat <- fit_cal_slope$vbeta
wald <- drop(betas %*% solve(vcov_mat) %*% betas)


betas_SCREAM_mm <- fit_cal_slope_SCREAM_mm$beta
vcov_mat_SCREAM_mm <- fit_cal_slope_SCREAM_mm$vbeta
wald_SCREAM_mm <- drop(betas_SCREAM_mm %*% solve(vcov_mat_SCREAM_mm) %*% betas_SCREAM_mm)

betas_SCREAM_nomm <- fit_cal_slope_SCREAM_nomm$beta
vcov_mat_SCREAM_nomm <- fit_cal_slope_SCREAM_nomm$vbeta
wald_SCREAM_nomm <- drop(betas_SCREAM_nomm %*% solve(vcov_mat_SCREAM_nomm) %*% betas_SCREAM_nomm)

betas_UKB_mm <- fit_cal_slope_UKB_mm$beta
vcov_mat_UKB_mm <- fit_cal_slope_UKB_mm$vbeta
wald_UKB_mm <- drop(betas_UKB_mm %*% solve(vcov_mat_UKB_mm) %*% betas_UKB_mm)

betas_UKB_nomm <- fit_cal_slope_UKB_nomm$beta
vcov_mat_UKB_nomm <- fit_cal_slope_UKB_nomm$vbeta
wald_UKB_nomm <- drop(betas_UKB_nomm %*% solve(vcov_mat_UKB_nomm) %*% betas_UKB_nomm)

pchisq(wald, df = 2, lower.tail = FALSE)
pchisq(wald_SCREAM_mm, df = 2, lower.tail = FALSE)
pchisq(wald_SCREAM_nomm, df = 2, lower.tail = FALSE)
pchisq(wald_UKB_mm, df = 2, lower.tail = FALSE)
pchisq(wald_UKB_nomm, df = 2, lower.tail = FALSE)

```
</details>

```{r, res_cal_intslope, echo=FALSE}
# Define a function to extract calibration results
extract_cal_results <- function(fit_cal_int, fit_cal_slope) {
  intercept <- with(
    summary(fit_cal_int)$mean,
    c(
      "estimate" = estimate,
      `2.5 %` = estimate - qnorm(0.975) * san.se,
      `97.5 %` = estimate + qnorm(0.975) * san.se
    )
  )
  
  slope <- with(
    summary(fit_cal_slope)$mean["cll_pred", ],
    c(
      "estimate" = 1 + estimate,
      `2.5 %` = 1 + (estimate - qnorm(0.975) * san.se),
      `97.5 %` = 1 + (estimate + qnorm(0.975) * san.se)
    )
  )
  
  return(c(intercept, slope))
}

# Extract results for all datasets
cal_results_pseudo <- extract_cal_results(fit_cal_int, fit_cal_slope)
cal_results_SCREAM_mm <- extract_cal_results(fit_cal_int_SCREAM_mm, fit_cal_slope_SCREAM_mm)
cal_results_SCREAM_nomm <- extract_cal_results(fit_cal_int_SCREAM_nomm, fit_cal_slope_SCREAM_nomm)
cal_results_UKB_mm <- extract_cal_results(fit_cal_int_UKB_mm, fit_cal_slope_UKB_mm)
cal_results_UKB_nomm <- extract_cal_results(fit_cal_int_UKB_nomm, fit_cal_slope_UKB_nomm)

# Combine results into a data frame
res_cal <- data.frame(
  Dataset = c("Pseudo", "SCREAM_mm", "SCREAM_nomm", "UKB_mm", "UKB_nomm"),
  Intercept = c(cal_results_pseudo[1], cal_results_SCREAM_mm[1], cal_results_SCREAM_nomm[1], cal_results_UKB_mm[1], cal_results_UKB_nomm[1]),
  Intercept_2.5 = c(cal_results_pseudo[2], cal_results_SCREAM_mm[2], cal_results_SCREAM_nomm[2], cal_results_UKB_mm[2], cal_results_UKB_nomm[2]),
  Intercept_97.5 = c(cal_results_pseudo[3], cal_results_SCREAM_mm[3], cal_results_SCREAM_nomm[3], cal_results_UKB_mm[3], cal_results_UKB_nomm[3]),
  Slope = c(cal_results_pseudo[4], cal_results_SCREAM_mm[4], cal_results_SCREAM_nomm[4], cal_results_UKB_mm[4], cal_results_UKB_nomm[4]),
  Slope_2.5 = c(cal_results_pseudo[5], cal_results_SCREAM_mm[5], cal_results_SCREAM_nomm[5], cal_results_UKB_mm[5], cal_results_UKB_nomm[5]),
  Slope_97.5 = c(cal_results_pseudo[6], cal_results_SCREAM_mm[6], cal_results_SCREAM_nomm[6], cal_results_UKB_mm[6], cal_results_UKB_nomm[6])
)


# Create the table
kable(res_cal) %>%
  kable_styling("striped", position = "center")


```

### 2.2 Discrimination

We here calculate

+ The 5-year C-index. More details are in the main manuscript and its references;
+ The 5-year time-dependent AUC. More details are in the manuscript and in its references;  
  + Plot time-dependent AUC over the time;
  
+ Royston-Sauerbrei D statistic and R<sup>2</sup><sub>D</sub> as a measure of prognostic separation;  

  
#### 2.2.1 C-index and time-dependent AUC

<details>
  <summary>Click to expand code</summary>
```{r, discrimination,warning=FALSE,message=FALSE,echo=TRUE}
# Models
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)

# useful objects
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# C-index
# Development set (Apparent validation)

C_rdata <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = rdata
)$AppCindex$CauseSpecificCox

# Validation set
C_vdata <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata
)$AppCindex$CauseSpecificCox

# SCREAM mm vs no mm
C_rdata_SCREAM_mm <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = rdata_mm
)$AppCindex$CauseSpecificCox

C_rdata_SCREAM_nomm <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = rdata_nomm
)$AppCindex$CauseSpecificCox

# UKB mm vs no mm
C_vdata_UKB_mm <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata_mm
)$AppCindex$CauseSpecificCox

C_vdata_UKB_nomm <- pec::cindex(
  object = fit_csh,
  formula = Hist(time, event) ~ 1,
  cause = primary_event,
  eval.times = horizon,
  data = vdata_nomm
)$AppCindex$CauseSpecificCox

# Bootstraping C-index to calculate the bootstrap percentile confidence intervals

B <- 100
set.seed(1234)
rboot <- bootstraps(rdata, times = B) # development - bootstrap
vboot <- bootstraps(vdata, times = B) # validation - bootstrap
rboot_SCREAM_mm <- bootstraps(rdata_mm, times = B) # SCREAM_mm - bootstrap
rboot_SCREAM_nomm <- bootstraps(rdata_nomm, times = B) # SCREAM_nomm - bootstrap
vboot_UKB_mm <- bootstraps(vdata_mm, times = B) # UKB_mm - bootstrap
vboot_UKB_nomm <- bootstraps(vdata_nomm, times = B) # UKB_nomm - bootstrap

C_boot <- function(split) {
  pec::cindex(
    object = fit_csh,
    formula = Hist(time, event) ~ 1,
    cause = primary_event,
    eval.times = horizon,
    data = analysis(split)
  )$AppCindex$CauseSpecificCox
}

bootstrap_cindex <- function(data, fit, primary_event, horizon, B) {
  rboot <- bootstraps(data, times = B)
  C_boot <- function(split) {
    pec::cindex(
      object = fit,
      formula = Hist(time, event) ~ 1,
      cause = primary_event,
      eval.times = horizon,
      data = analysis(split)
    )$AppCindex$CauseSpecificCox
  }
  rboot <- rboot %>% mutate(C_boot = map_dbl(splits, C_boot))
  ci <- quantile(rboot$C_boot, probs = c(0.025, 0.975), na.rm = TRUE)
  list(C_boot = rboot$C_boot, ci = ci)
}
boot_dev <- bootstrap_cindex(rdata, fit_csh, primary_event, horizon, B)
boot_val <- bootstrap_cindex(vdata, fit_csh, primary_event, horizon, B)
boot_SCREAM_mm <- bootstrap_cindex(rdata_mm, fit_csh, primary_event, horizon, B)
boot_SCREAM_nomm <- bootstrap_cindex(rdata_nomm, fit_csh, primary_event, horizon, B)
boot_UKB_mm <- bootstrap_cindex(vdata_mm, fit_csh, primary_event, horizon, B)
boot_UKB_nomm <- bootstrap_cindex(vdata_nomm, fit_csh, primary_event, horizon, B)

# Run time-dependent AUC in the bootstrapped development and validation data
# to calculate the non-parametric CI through percentile bootstrap
rboot <- rboot |> mutate(
  C_rboot = map_dbl(splits, C_boot),
)
vboot <- vboot |> mutate(
  C_vboot = map_dbl(splits, C_boot),
)
rboot_SCREAM_mm <- rboot_SCREAM_mm |> mutate(
  C_rboot_SCREAM_mm = map_dbl(splits, C_boot),
)
rboot_SCREAM_mm <- rboot_SCREAM_mm |> mutate(
  C_rboot_SCREAM_mm = map_dbl(splits, C_boot),
)
vboot_UKB_mm <- vboot_UKB_mm |> mutate(
  C_vboot_UKB_mm = map_dbl(splits, C_boot),
)
vboot_UKB_mm <- vboot_UKB_mm |> mutate(
  C_vboot_UKB_mm = map_dbl(splits, C_boot),
)

# Time-dependent AUC ---------

# Development data
score_rdata <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

# Validation data
score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

# SCREAM mm vs nomm
score_rdata_SCREAM_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_mm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

score_rdata_SCREAM_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_nomm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

# UKB mm vs nomm
score_vdata_UKB_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_mm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)

score_vdata_UKB_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_nomm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc"),
  cause = primary_event,
  plots = "calibration"
)
```
</details>

```{r, res_disc, fig.align='center',echo=FALSE}


cindex_results <- data.frame(
  Group = c("Development", "Validation", "SCREAM mm", "SCREAM nomm", "UKB mm", "UKB nomm"),
  Cindex = c(C_rdata, C_vdata, C_rdata_SCREAM_mm, C_rdata_SCREAM_nomm, C_vdata_UKB_mm, C_vdata_UKB_nomm),
  `2.5%` = c(boot_dev$ci[1], boot_val$ci[1], boot_SCREAM_mm$ci[1], boot_SCREAM_nomm$ci[1], boot_UKB_mm$ci[1], boot_UKB_nomm$ci[1]),
  `97.5%` = c(boot_dev$ci[2], boot_val$ci[2], boot_SCREAM_mm$ci[2], boot_SCREAM_nomm$ci[2], boot_UKB_mm$ci[2], boot_UKB_nomm$ci[2])
)

# Print the table
kable(cindex_results, col.names = c("Group", "Cindex", "2.5%", "97.5%")) %>%
  kable_styling("striped", position = "center")


  # AUC and Confidence Intervals

calculate_auc <- function(data, fit, primary_event, horizon) {
  score_obj <- Score(
    list("csh" = fit),
    formula = Hist(time, event) ~ 1,
    cens.model = "km",
    data = data,
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc"),
    cause = primary_event,
    plots = "calibration"
  )
  
  auc <- score_obj$AUC$score$AUC
  se <- score_obj$AUC$score$se
  list(auc = auc, se = se)
}

# Calculate AUC for all groups
score_rdata <- calculate_auc(rdata, fit_csh, primary_event, horizon)
score_vdata <- calculate_auc(vdata, fit_csh, primary_event, horizon)
score_rdata_SCREAM_mm <- calculate_auc(rdata_mm, fit_csh, primary_event, horizon)
score_rdata_SCREAM_nomm <- calculate_auc(rdata_nomm, fit_csh, primary_event, horizon)
score_vdata_UKB_mm <- calculate_auc(vdata_mm, fit_csh, primary_event, horizon)
score_vdata_UKB_nomm <- calculate_auc(vdata_nomm, fit_csh, primary_event, horizon)

alpha <- 0.05
k <- 2

AUC_results <- data.frame(
  Group = c("Development", "Validation", "SCREAM mm", "SCREAM nomm", "UKB mm", "UKB nomm"),
  AUC = c(score_rdata$auc, score_vdata$auc, score_rdata_SCREAM_mm$auc, score_rdata_SCREAM_nomm$auc, score_vdata_UKB_mm$auc, score_vdata_UKB_nomm$auc),
  `2.5%` = c(score_rdata$auc - qnorm(1 - alpha / 2) * score_rdata$se, score_vdata$auc - qnorm(1 - alpha / 2) * score_vdata$se, score_rdata_SCREAM_mm$auc - qnorm(1 - alpha / 2) * score_rdata_SCREAM_mm$se, score_rdata_SCREAM_nomm$auc - qnorm(1 - alpha / 2) * score_rdata_SCREAM_nomm$se, score_vdata_UKB_mm$auc - qnorm(1 - alpha / 2) * score_vdata_UKB_mm$se, score_vdata_UKB_nomm$auc - qnorm(1 - alpha / 2) * score_vdata_UKB_nomm$se),
  `97.5%` = c(score_rdata$auc + qnorm(1 - alpha / 2) * score_rdata$se, score_vdata$auc + qnorm(1 - alpha / 2) * score_vdata$se, score_rdata_SCREAM_mm$auc + qnorm(1 - alpha / 2) * score_rdata_SCREAM_mm$se, score_rdata_SCREAM_nomm$auc + qnorm(1 - alpha / 2) * score_rdata_SCREAM_nomm$se, score_vdata_UKB_mm$auc + qnorm(1 - alpha / 2) * score_vdata_UKB_mm$se, score_vdata_UKB_nomm$auc + qnorm(1 - alpha / 2) * score_vdata_UKB_nomm$se)
)

kable(AUC_results, col.names = c("Group", "AUC", "2.5%", "97.5%")) %>%
  kable_styling("striped", position = "center")
```


The time-dependent AUC at 5 years was 0.92 (95%CI 0.90-0.94) and c-index was 0.92 (95% CI 0.90-0.93) in the validation (UK Biobank) set.

#### 2.2.2 Plot Area under the curve(s) over the time
We plot the time-dependent AUCs over the follow-up time using development and validation data. 

<details>
  <summary>Click to expand code</summary>
```{r, plot_AUCs, warning=FALSE,echo=TRUE, eval=FALSE}

# Models --------------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)
primary_event <- 1 # Set to 2 if cause 2 was of interest

# AUCs development data
AUC_rdata <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata,
  conf.int = TRUE,
  times = seq(0.4, 5.1, 0.2),
  metrics = c("auc"),
  cause = primary_event
)

# AUCs validation data
AUC_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = seq(0.5, 5.01, 0.1),
  metrics = c("auc"),
  cause = primary_event
)

# AUCs for SCREAM and UKB data
AUC_rdata_SCREAM_mm <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_mm,
  conf.int = TRUE,
  times = seq(0.4, 5.1, 0.2),
  metrics = c("auc"),
  cause = primary_event
)

AUC_rdata_SCREAM_nomm <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_nomm,
  conf.int = TRUE,
  times = seq(0.4, 5.1, 0.2),
  metrics = c("auc"),
  cause = primary_event
)

AUC_vdata_UKB_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_mm,
  conf.int = TRUE,
  times = seq(0.5, 5.01, 0.1),
  metrics = c("auc"),
  cause = primary_event
)

AUC_vdata_UKB_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_nomm,
  conf.int = TRUE,
  times = seq(0.5, 5.01, 0.1),
  metrics = c("auc"),
  cause = primary_event
)

# Plot
par(las = 1, xaxs = "i", yaxs = "i")
oldpar <- par(mfrow = c(1, 2))
plot(AUC_rdata$times,
  AUC_rdata$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_rdata$times,
  rev(AUC_rdata$times)
),
c(
  AUC_rdata$AUC$score$lower,
  rev(AUC_rdata$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_rdata$times,
  AUC_rdata$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("Development data", adj = 0)

# Validation data
plot(AUC_vdata$times,
  AUC_vdata$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_vdata$times,
  rev(AUC_vdata$times)
),
c(
  AUC_vdata$AUC$score$lower,
  rev(AUC_vdata$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_vdata$times,
  AUC_vdata$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("Validation data", adj = 0)

# SCREAM mm vs nomm data
plot(AUC_rdata_SCREAM_mm$times,
  AUC_rdata_SCREAM_mm$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_rdata_SCREAM_mm$times,
  rev(AUC_rdata_SCREAM_mm$times)
),
c(
  AUC_rdata_SCREAM_mm$AUC$score$lower,
  rev(AUC_rdata_SCREAM_mm$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_rdata_SCREAM_mm$times,
  AUC_rdata_SCREAM_mm$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("SCREAM multimorbidity cohort", adj = 0)

plot(AUC_rdata_SCREAM_nomm$times,
  AUC_rdata_SCREAM_nomm$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_rdata_SCREAM_nomm$times,
  rev(AUC_rdata_SCREAM_nomm$times)
),
c(
  AUC_rdata_SCREAM_nomm$AUC$score$lower,
  rev(AUC_rdata_SCREAM_nomm$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_rdata_SCREAM_nomm$times,
  AUC_rdata_SCREAM_nomm$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("SCREAM no multimorbidity cohort", adj = 0)

# UKB mm vs nomm data
plot(AUC_vdata_UKB_mm$times,
  AUC_vdata_UKB_mm$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_vdata_UKB_mm$times,
  rev(AUC_vdata_UKB_mm$times)
),
c(
  AUC_vdata_UKB_mm$AUC$score$lower,
  rev(AUC_vdata_UKB_mm$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_vdata_UKB_mm$times,
  AUC_vdata_UKB_mm$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("UKB multimorbidity cohort", adj = 0)

plot(AUC_vdata_UKB_nomm$times,
  AUC_vdata_UKB_nomm$AUC$score$AUC,
  type = "l",
  bty = "n",
  xlim = c(0, 5),
  ylim = c(.5, 1),
  lwd = 2,
  xlab = "Time (years)",
  ylab = "AUC",
  lty = 1
)
polygon(c(
  AUC_vdata_UKB_nomm$times,
  rev(AUC_vdata_UKB_nomm$times)
),
c(
  AUC_vdata_UKB_nomm$AUC$score$lower,
  rev(AUC_vdata_UKB_nomm$AUC$score$upper)
),
col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
border = FALSE
)
lines(AUC_vdata_UKB_nomm$times,
  AUC_vdata_UKB_nomm$AUC$score$AUC,
  col = "black",
  lwd = 2,
  lty = 2
)
title("UKB no multimorbidity cohort", adj = 0)
```
</details>

```{r, plot_AUCs, warning=FALSE,echo=FALSE, eval=TRUE}
```



### 2.3 Overall prediction error

We calculate the Brier Score, and the scaled Brier scale (also indicated by scaled Brier score or IPA) and the corresponding confidence intervals.

Some confidence intervals are calculated using the bootstrap percentile method.
```{r, bootstrap,warning=FALSE}
# Bootstrapping data
set.seed(20201214)
B <- 500 # number of bootstrap samples
rboot <- bootstraps(rdata, times = B)
vboot <- bootstraps(vdata, times = B)
rboot_SCREAM_mm <- bootstraps(rdata_mm, times = B) 
rboot_SCREAM_nomm <- bootstraps(rdata_nomm, times = B) 
vboot_UKB_mm <- bootstraps(vdata_mm, times = B)
vboot_UKB_nomm <- bootstraps(vdata_nomm, times = B)

```

<details>
  <summary>Click to expand code</summary>
```{r, overall, warning=FALSE,  fig.align='center', message=FALSE, echo=TRUE}

# Models -------------------
fit_csh <- CSC(Hist(time, event) ~
age + sexnumber + egfrcr09_race_allnonblack + lnacr2,
data = rdata,
fitter = "cph"
)
fit_csc1 <- fit_csh$models$`Cause 1`
fit_csc2 <- fit_csh$models$`Cause 2`

# Overall performance measures ----------------
primary_event <- 1 # Set to 2 if cause 2 was of interest
horizon <- 5 # Set time horizon for prediction (here 5 years)

# Development data
score_rdata <- Score(
  list("csh_development" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# Validation data
score_vdata <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# SCREAM multimorbidity
score_rdata_SCREAM_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_mm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# SCREAM no multimorbidity
score_rdata_SCREAM_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = rdata_nomm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# UKB multimorbidity
score_vdata_UKB_mm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_mm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)

# UKB no multimorbidity
score_vdata_UKB_nomm <- Score(
  list("csh_validation" = fit_csh),
  formula = Hist(time, event) ~ 1,
  cens.model = "km",
  data = vdata_nomm,
  conf.int = TRUE,
  times = horizon,
  metrics = c("auc", "brier"),
  summary = c("ipa"),
  cause = primary_event,
  plots = "calibration"
)


# Bootstrap ------
# Functions to expand data and calculate Brier, IPA and AUC in bootstrap
# samples.
# For Brier and AUC, bootstrap should be computationally faster when
# data has more than 2000 rows (see ?riskRegression::Score).
# Our data has 1000 row so we will need only bootstrap to calculate
# confidence intervals of the scaled Brier (IPA) since
# it is not provided by riskRegression::Score() function.


# Score functions in any bootstrap data
score_boot <- function(split) {
  Score(
    list("csh_validation" = fit_csh),
    formula = Hist(time, event) ~ 1,
    cens.model = "km",
    data = analysis(split),
    conf.int = TRUE,
    times = horizon,
    metrics = c("auc", "brier"),
    summary = c("ipa"),
    cause = primary_event,
    plots = "calibration"
  )
}

# Development data
rboot <- rboot |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# Validation data
vboot <- vboot |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# SCREAM mm
rboot_SCREAM_mm <- rboot_SCREAM_mm |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# SCREAM no mm
rboot_SCREAM_nomm <- rboot_SCREAM_nomm |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# UKB mm
vboot_UKB_mm <- vboot_UKB_mm |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
# UKB no mm
vboot_UKB_nomm <- vboot_UKB_nomm |> mutate(
  score = map(splits, score_boot),
  scaled_brier = map_dbl(score, function(x) {
    x$Brier$score[model == "csh_validation"]$IPA
  })
)
```
</details>

```{r, overall, warning=FALSE,echo=FALSE, eval=TRUE}
```

```{r, res_ov, fig.align='center',echo=FALSE}
# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov_csh <- matrix(unlist(c(
  # Brier score - apparent validation
  score_rdata$Brier$score[model == "csh_development"]$Brier,
  score_rdata$Brier$score[model == "csh_development"]$lower,
  score_rdata$Brier$score[model == "csh_development"]$upper,

  # Brier score - external validation
  score_vdata$Brier$score[model == "csh_validation"]$Brier,
  score_vdata$Brier$score[model == "csh_validation"]$lower,
  score_vdata$Brier$score[model == "csh_validation"]$upper,
  
  # Brier score - SCREAM mm
  score_rdata_SCREAM_mm$Brier$score[model == "csh_validation"]$Brier,
  score_rdata_SCREAM_mm$Brier$score[model == "csh_validation"]$lower,
  score_rdata_SCREAM_mm$Brier$score[model == "csh_validation"]$upper,
 
  # Brier score - SCREAM no mm
  score_rdata_SCREAM_nomm$Brier$score[model == "csh_validation"]$Brier,
  score_rdata_SCREAM_nomm$Brier$score[model == "csh_validation"]$lower,
  score_rdata_SCREAM_nomm$Brier$score[model == "csh_validation"]$upper,
   
  # Brier score - UKB mm
  score_vdata_UKB_mm$Brier$score[model == "csh_validation"]$Brier,
  score_vdata_UKB_mm$Brier$score[model == "csh_validation"]$lower,
  score_vdata_UKB_mm$Brier$score[model == "csh_validation"]$upper,
  
  # Brier score - UKB no mm
  score_vdata_UKB_nomm$Brier$score[model == "csh_validation"]$Brier,
  score_vdata_UKB_nomm$Brier$score[model == "csh_validation"]$lower,
  score_vdata_UKB_nomm$Brier$score[model == "csh_validation"]$upper,

  # IPA - apparent validation
  score_rdata$Brier$score[model == "csh_development"]$IPA,
  quantile(rboot$scaled_brier, probs = alpha / 2),
  quantile(rboot$scaled_brier, probs = 1 - alpha / 2),

  # IPA - external validation
  score_vdata$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot$scaled_brier, probs = alpha / 2),
  quantile(vboot$scaled_brier, probs = 1 - alpha / 2),
  
   # IPA - SCREAM mm
  score_rdata_SCREAM_mm$Brier$score[model == "csh_validation"]$IPA,
  quantile(rboot_SCREAM_mm$scaled_brier, probs = alpha / 2),
  quantile(rboot_SCREAM_mm$scaled_brier, probs = 1 - alpha / 2),
  
  # IPA - SCREAM no mm
  score_rdata_SCREAM_nomm$Brier$score[model == "csh_validation"]$IPA,
  quantile(rboot_SCREAM_nomm$scaled_brier, probs = alpha / 2),
  quantile(rboot_SCREAM_nomm$scaled_brier, probs = 1 - alpha / 2),
  
   # IPA - UKB mm
  score_vdata_UKB_mm$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot_UKB_mm$scaled_brier, probs = alpha / 2),
  quantile(vboot_UKB_mm$scaled_brier, probs = 1 - alpha / 2),
  
  # IPA - UKB no mm
  score_vdata_UKB_nomm$Brier$score[model == "csh_validation"]$IPA,
  quantile(vboot_UKB_nomm$scaled_brier, probs = alpha / 2),
  quantile(vboot_UKB_nomm$scaled_brier, probs = 1 - alpha / 2)
)),
nrow = 2, ncol = 18,
byrow = T,
dimnames =
  list(
    c("Brier", "scaled Brier"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"),6)
  )
)



kable(res_ov_csh) |>
  kable_styling("striped", position = "center")  |>
  add_header_above(c(
    " " = 1,
    "Development data" = 3,
    "Validation data" = 3,
    "SCREAM multimorbidity" = 3,
    "SCREAM no multimorbidity" = 3,
    "UKB multimorbidity" = 3,
    "UKB no multimorbidity" = 3
  ))
```

```{r details for reporting full model formula}
model_info2 <- list(
  "coefficients" = stats::coef(fit_csh),
  "baseline_hazards" = lapply(fit_csh$models, function(mod) survival::basehaz(mod, centered = FALSE)),
  "model_terms" = lapply(fit_csh$models, function(mod) mod[["terms"]])
)

model_info2 

# Can be saved to then be shared as:
saveRDS(model_info2, file = "model_info2.rds")
```

##Compare original KFRE risk referral thresholds with competing risk referral thresholds

```{r}
# Re-import data ------------------
load(file="~/R working directory/SCREAM/linear_predictor.RData")
SCREAM_plus_CSCrisk <- linear_predictor
UKB_plus_CSCrisk <- readRDS("~/R working directory/UKB/new_ukb.rds")

UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>% mutate(cardio_mm = ifelse(card_met_mmcount_cat==0|card_met_mmcount_cat==1, 0, 1))

UKB_plus_CSCrisk$event <- ifelse(UKB_plus_CSCrisk$death==1, 2, 0)
UKB_plus_CSCrisk$event <- ifelse(UKB_plus_CSCrisk$outcome==1, 1, UKB_plus_CSCrisk$event)
UKB_plus_CSCrisk = UKB_plus_CSCrisk %>% 
  mutate(time = (futime_months/12)) %>% 
  mutate(status= case_when(
    event==2 ~ "death",
    event==1 ~"KF",
    event==0 ~ "cens")) %>% 
  mutate(acr2 = ifelse(acr2 == 0, 0.00001, acr2)) %>% 
  mutate(lnacr2 = (log(acr2)))

SCREAM_plus_CSCrisk$event <- ifelse(SCREAM_plus_CSCrisk$death==1, 2, 0)
SCREAM_plus_CSCrisk$event <- ifelse(SCREAM_plus_CSCrisk$outcome==1, 1, SCREAM_plus_CSCrisk$event)
SCREAM_plus_CSCrisk = SCREAM_plus_CSCrisk %>% 
  mutate(time = (futime_months/12)) %>% 
  mutate(status= case_when(
    event==2 ~ "death",
    event==1 ~"KF",
    event==0 ~ "cens")) %>% 
  mutate(acr2 = ifelse(acr2 == 0, 0.00001, acr2)) %>% 
  mutate(lnacr2 = (log(acr2)))

UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>%
  mutate(status = factor(case_when(
    event == 2 ~ "death",
    event == 1 ~ "KF",
    event == 0 ~ "cens"
  ), levels = c("death", "KF", "cens")))

SCREAM_plus_CSCrisk <- SCREAM_plus_CSCrisk %>%
  mutate(status = factor(case_when(
    event == 2 ~ "death",
    event == 1 ~ "KF",
    event == 0 ~ "cens"
  ), levels = c("death", "KF", "cens")))

SCREAM_plus_CSCrisk <- SCREAM_plus_CSCrisk %>%
  mutate(sex = factor(case_when(
    female == 0 ~ "female",
    female == 1 ~ "male"
  ), levels = c("female", "male")))


UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>%
  mutate(sex = factor(case_when(
    sexnumber == 0 ~ "female",
    sexnumber == 1 ~ "male"
  ), levels = c("female", "male")))

SCREAM_plus_CSCrisk <- SCREAM_plus_CSCrisk %>% 
  rename_at('lopnr', ~'eid') %>% 
  rename_at('sex', ~'sexnumber')
UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>% 
  select(-sexnumber) %>% 
  rename_at('sex', ~'sexnumber')

UKB_plus_CSCrisk_mm <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$mmcount_nockd >1)
UKB_plus_CSCrisk_nomm <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$mmcount_nockd<=1)
SCREAM_plus_CSCrisk_mm <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$mmcount_nockd >1)
SCREAM_plus_CSCrisk_nomm <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$mmcount_nockd<=1)


# Calculate predicted probabilities
SCREAM_plus_CSCrisk$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = SCREAM_plus_CSCrisk,
  times = horizon
)

UKB_plus_CSCrisk$pred <- predictRisk(
  fit_csh,
  cause = 1,
  newdata = UKB_plus_CSCrisk,
  times = horizon
)

#make age categories
SCREAM_plus_CSCrisk <- SCREAM_plus_CSCrisk %>% 
  mutate(age_cat = cut(age, 
      breaks = c(18, 44.9, 64.9, 74.9, 107), 
      include.lowest = TRUE)) %>%
  mutate(age_cat = fct_recode(age_cat,
      "18to44" = "[18,44.9]",
      "45to64" = "(44.9,64.9]",
      "65to74" = "(64.9,74.9]",
      "75plus" = "(74.9,107]"))

UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>% 
  mutate(age_cat = cut(age, 
      breaks = c(18, 44.9, 64.9, 74.9), 
      include.lowest = TRUE)) %>%
  mutate(age_cat = fct_recode(age_cat,
      "18to44" = "[18,44.9]",
      "45to64" = "(44.9,64.9]",
      "65to74" = "(64.9,74.9]"))

UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>% 
  mutate(mmcount_nockd_cat = fct_recode(mmcount_nockd_cat,
                                        "0" = "[0,0.9]",
                                        "1" = "(0.9,1.9]",
                                        "2" = "(1.9,2.9]",
                                        "3" = "(2.9,3.9]",
                                        "4" = "(3.9,4.9]",
                                        "5 or more" = "(4.9,13]"))

#Create risk groups for competing risk model (pred)
cutoffs <- c(-Inf, 0.03, 0.05, 0.15, 0.25, 0.50, Inf)

SCREAM_plus_CSCrisk <- SCREAM_plus_CSCrisk %>% 
  mutate(risk_group_CSC = cut(pred, breaks = cutoffs, 
                                  include.lowest = TRUE,
                                  labels = c("<3%", "3-<5%", "5-<15%", "15-<25%", "25-<50%", ">=50%")))


UKB_plus_CSCrisk <- UKB_plus_CSCrisk %>% 
  mutate(risk_group_CSC = cut(pred, breaks = cutoffs, 
                                  include.lowest = TRUE,
                                  labels = c("<3%", "3-<5%", "5-<15%", "15-<25%", "25-<50%", ">=50%")))


UKB_KFRE_GT3 <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$KFRE_egfrcr09_race_allnonblack1>=0.03)
UKB_KFRE_GT5 <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$KFRE_egfrcr09_race_allnonblack1>=0.05)
UKB_CSC_GT3 <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$pred>=0.03)
UKB_CSC_GT5 <- subset(UKB_plus_CSCrisk, UKB_plus_CSCrisk$pred>=0.05)
SCREAM_KFRE_GT3 <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$KFRE_egfrcr09_race_allnonblack1>=0.03)
SCREAM_KFRE_GT5 <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$KFRE_egfrcr09_race_allnonblack1>=0.05)
SCREAM_CSC_GT3 <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$pred>=0.03)
SCREAM_CSC_GT5 <- subset(SCREAM_plus_CSCrisk, SCREAM_plus_CSCrisk$pred>=0.05)


```

### Descriptive statistics

```{r, import}
SCREAM_KFRE_GT5 <- SCREAM_KFRE_GT5[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
SCREAM_CSC_GT5 <- SCREAM_CSC_GT5[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
UKB_KFRE_GT5 <- UKB_KFRE_GT5[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
UKB_CSC_GT5 <- UKB_CSC_GT5[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]

SCREAM_KFRE_GT3 <- SCREAM_KFRE_GT3[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
SCREAM_CSC_GT3 <- SCREAM_CSC_GT3[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
UKB_KFRE_GT3 <- UKB_KFRE_GT3[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]
UKB_CSC_GT3 <- UKB_CSC_GT3[, c("eid", "age", "sexnumber", "egfrcr09_race_allnonblack", "lnacr2", "acr2", "age_cat", "mmcount_nockd_cat", "cardio_mm", "mixedLTC", "complexLTC", "outcome_5yrs", "death_5yrs")]



UKB_KFRE_GT5$dt <- 1
UKB_CSC_GT5$dt <- 2
SCREAM_KFRE_GT5$dt <- 3
SCREAM_CSC_GT5$dt <- 4
cdata <- rbind(UKB_KFRE_GT5, UKB_CSC_GT5, SCREAM_KFRE_GT5, SCREAM_CSC_GT5)
cdata$dt <- factor(
  cdata$dt,
  levels = c(1, 2, 3, 4),
  labels = c("UK Biobank KFRE", "UK Bioank Competing Risk model", "SCREAM KFRE", "SCREAM Competing Risk model")
)
label(cdata$age) <- "Age"
label(cdata$sexnumber) <- "Sex"
label(cdata$egfrcr09_race_allnonblack) <- "eGFR creatinine"
label(cdata$acr2) <- "uACR"
units(cdata$age) <- "years"
units(cdata$egfrcr09_race_allnonblack) <- "ml/min/1.73m2"
units(cdata$acr2) <- "mg/mmol"
```


```{r tab1_bis}
gtsummary::tbl_summary(
  data = cdata |> select(-eid),
  label = list(age ~ "Age (years)", egfrcr09_race_allnonblack ~ "eGFR creatinine (ml/min/1.73m2)", acr2 ~"uACR (mg/mmol)"),
  by = "dt",
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
) |>
  gtsummary::as_kable_extra() |>
  kableExtra::kable_styling("striped")
```



  
